---
title: "Solve-A-Thon 2026: NC DHHS Federal Grant Opportunity Analysis"
author: "Kent Lee"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    smooth-scroll: true
    page-layout: full
execute:
  warning: true
  message: true
  error: true
  echo: false
params:
  project_dir: "/Users/kentalee/Documents/KALProjects/UNCProjects/SolveAThon_2026"
  data_dir: "data"
  results_dir: "outs"
  cache_dir: "cache"
  grants_file: "grants-search-202608182008.csv"
  census_api_key: ""
  ref_date: "2026-09-30"
  min_runway_days: 45
  award_floor_usd: 250000
  target_agencies: "DHHS"
  nc_fips: "37"
  map_geography: "county"
  boundary_year: 2023
  date_min: "2000-01-01"
  date_max: "2035-12-31"
---


<style>
:root {
  --blue:#2166AC; --blue-light:#D9EAF7; --purple:#756BB1;
  --orange:#E08214; --red:#B2182B; --green:#1B7837;
  --gray:#667085; --dark:#1D2939; --light:#F4F6F8;
}
body { color:var(--dark); }
h1,h2 { border-bottom:2px solid var(--blue-light); padding-bottom:.35rem; margin-top:2rem; }
h3 { margin-top:1.5rem; }
.kpi-grid { display:grid; grid-template-columns:repeat(auto-fit,minmax(180px,1fr)); gap:14px; margin:18px 0 28px 0; }
.kpi-card { background:white; border:1px solid #E4E7EC; border-left:5px solid var(--blue); border-radius:8px; padding:16px; min-height:100px; box-shadow:0 2px 7px rgba(16,24,40,.06); }
.kpi-card.warning { border-left-color:var(--orange); }
.kpi-card.danger  { border-left-color:var(--red); }
.kpi-card.success { border-left-color:var(--green); }
.kpi-label { color:var(--gray); font-size:.82rem; text-transform:uppercase; letter-spacing:.04em; }
.kpi-value { color:var(--dark); font-size:1.8rem; font-weight:700; margin-top:5px; }
.kpi-note  { color:var(--gray); font-size:.78rem; margin-top:5px; }
.callout { background:#F8FAFC; border-left:5px solid var(--blue); padding:14px 18px; margin:18px 0; border-radius:4px; }
.callout-warning { border-left-color:var(--orange); background:#FFFAEB; }
.callout-green   { border-left-color:var(--green);  background:#EDF7ED; }
.data-source-box { background:#F0F7F0; border-left:5px solid var(--green); padding:12px 16px; margin:14px 0; border-radius:4px; font-size:.88em; }
.dataTables_wrapper { font-size:.88em; margin-top:15px; margin-bottom:30px; }
table.dataTable thead th { background-color:#EEF4FA; }
pre { border-radius:6px; }
</style>


```{r setup}
# Diagnostic mode: continue rendering after chunk errors so one render exposes
# the full error cascade instead of stopping at the first failure.
diagnostic_errors <- list()
diagnostic_warnings <- list()

knitr::knit_hooks$set(
  error = function(x, options) {
    diagnostic_errors[[length(diagnostic_errors) + 1L]] <<- list(
      label = options$label %||% "unknown-chunk",
      message = x
    )
    x
  },
  warning = function(x, options) {
    diagnostic_warnings[[length(diagnostic_warnings) + 1L]] <<- list(
      label = options$label %||% "unknown-chunk",
      message = x
    )
    x
  }
)

`%||%` <- function(x, y) if (is.null(x) || length(x) == 0L) y else x

knitr::opts_chunk$set(error=TRUE, 
  echo=FALSE, eval=TRUE, message=TRUE, warning=TRUE, error=TRUE,
  fig.align="center", dpi=150, out.width="100%"
)
project_dir <- params$project_dir
data_dir    <- file.path(project_dir, params$data_dir)
results_dir <- file.path(project_dir, params$results_dir)
cache_dir   <- file.path(data_dir,    params$cache_dir)
knitr::opts_knit$set(root.dir = project_dir)
for (d in c(results_dir, cache_dir,
            file.path(cache_dir, "api"),
            file.path(data_dir, "raw", "boundaries"),
            file.path(data_dir, "raw", "acs"),
            file.path(data_dir, "raw", "hrsa"),
            file.path(data_dir, "raw", "cdc_places"),
            file.path(data_dir, "raw", "cdc_wonder"),
            file.path(data_dir, "raw", "fema"),
            file.path(data_dir, "raw", "usaspending"),
            file.path(data_dir, "raw", "samhsa"),
            file.path(data_dir, "processed"),
            file.path(project_dir, "docs"),
            file.path(project_dir, "outputs", "tables"))) {
  dir.create(d, showWarnings=FALSE, recursive=TRUE)
}
options(bitmapType="cairo", tigris_use_cache=TRUE)
```


```{r libraries}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(lubridate); library(janitor); library(purrr)
  library(ggplot2); library(scales); library(DT)
  library(plotly); library(forcats); library(htmltools)
  library(sf); library(tigris); library(tidycensus)
  library(httr2); library(jsonlite)
  library(sf)
  library(tigris)
  library(tidycensus)
  library(leaflet)
  library(htmlwidgets)
  library(httr2)
  library(jsonlite)
})
# arrow: optional — used in Phase 4 for parquet output
arrow_available <- requireNamespace("arrow", quietly = TRUE)
if (arrow_available) suppressPackageStartupMessages(library(arrow))
```


```{r config}
ref_date_input                  <- params$ref_date
min_runway_days                 <- params$min_runway_days
award_floor_usd                 <- params$award_floor_usd
target_agencies                 <- params$target_agencies
include_forecasts               <- FALSE
exclude_research_mechanisms     <- TRUE
exclude_restricted_competitions <- TRUE
keep_forecasts                  <- include_forecasts


ref_date       <- ymd(ref_date_input)
earliest_close <- ref_date + days(min_runway_days)

# Plotting limits prevent extreme dates and award values from distorting figures.
# The 400-day runway limit follows the project's documented application-window
# screening range. It affects visualization only, not the underlying data.
plot_max_runway_days <- 400L

# Use a data-driven award cap for visualizations.
# The full data remain available in tables and screening results.
plot_award_quantile <- 0.99

nc_fips          <- params$nc_fips
map_geography    <- tolower(str_squish(params$map_geography))
boundary_year    <- as.integer(params$boundary_year)
date_min         <- ymd(params$date_min)
date_max         <- ymd(params$date_max)
peer_fips <- c("13","51","45","47")
peer_abbr <- c("GA","VA","SC","TN")


if (nchar(params$census_api_key) > 0)
  tidycensus::census_api_key(params$census_api_key, install=FALSE)


pal <- list(
  blue="#2166AC", blue_light="#D9EAF7", purple="#756BB1",
  orange="#E08214", red="#B2182B", green="#1B7837",
  gray="#667085", dark="#1D2939", light="#F4F6F8"
)


theme_g <- function(base_size=12) {
  theme_minimal(base_size=base_size) +
    theme(
      plot.title    = element_text(face="bold", size=base_size+3, color=pal$dark),
      plot.subtitle = element_text(color=pal$gray, margin=margin(b=12)),
      plot.caption  = element_text(color=pal$gray, size=base_size-2, hjust=0),
      axis.title    = element_text(color=pal$dark),
      axis.text     = element_text(color="#475467"),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.margin   = margin(12,20,12,12)
    )
}
theme_set(theme_g())
```


```{r helper-functions, include=FALSE}
is_unset <- function(x) {
  is.null(x) || length(x) == 0L || all(is.na(x))
}

parse_logical_safe <- function(x) {
  y <- str_to_lower(str_squish(as.character(x)))
  out <- rep(NA, length(y))
  out[y %in% c("true", "t", "yes", "y", "1")] <- TRUE
  out[y %in% c("false", "f", "no", "n", "0")] <- FALSE
  out
}

safe_false <- function(x) {
  dplyr::coalesce(parse_logical_safe(x), FALSE)
}


get_cached <- function(path, fetch_fn) {
  if (file.exists(path)) {
    cached <- tryCatch(
      readRDS(path),
      error = function(e) {
        message("Cached file unreadable; refreshing: ", path)
        NULL
      }
    )
    if (!is.null(cached)) return(cached)
  }

  result <- tryCatch(
    fetch_fn(),
    error = function(e) {
      message("Fetch failed: ", conditionMessage(e))
      NULL
    }
  )

  if (!is.null(result)) {
    dir.create(dirname(path), showWarnings = FALSE, recursive = TRUE)
    saveRDS(result, path)
  }

  result
}
```


```{r load-and-clean}
input_file <- file.path(data_dir, "raw", "grants", params$grants_file)
if (!file.exists(input_file)) input_file <- file.path(data_dir, params$grants_file)
if (!file.exists(input_file)) stop("Grants file not found: ", input_file)


grants_raw <- readr::read_csv(
  input_file,
  show_col_types = FALSE,
  na = c("", "NA", "N/A", "NULL")
) |>
  janitor::clean_names()

# Normalize optional Grants.gov fields so the rest of the report can run even
# when a snapshot omits a column.
required_character_cols <- c(
  "post_date", "close_date", "archive_date",
  "forecasted_post_date", "forecasted_close_date",
  "forecasted_award_date", "forecasted_project_start_date",
  "opportunity_title", "agency_name", "top_level_agency_name",
  "summary_description", "applicant_types", "funding_instruments",
  "funding_categories", "applicant_eligibility_description",
  "category", "category_explanation", "opportunity_status",
  "opportunity_number", "opportunity_assistance_listings",
  "additional_info_url", "cost_sharing_or_matching_requirement"
)

for (nm in setdiff(required_character_cols, names(grants_raw))) {
  grants_raw[[nm]] <- NA_character_
}

if (!"is_forecast" %in% names(grants_raw)) {
  grants_raw$is_forecast <- FALSE
}

parse_date_safe <- function(x) {
  x <- trimws(as.character(x))
  x[x %in% c("", "NA", "N/A", "NULL", "null", "NaN")] <- NA_character_
  x <- sub("T.*$", "", x, perl = TRUE)
  x <- sub(" .*$", "", x, perl = TRUE)
  result <- suppressWarnings(as.Date(x))
  for (fmt in c("%m/%d/%Y", "%m/%d/%y", "%Y/%m/%d", "%d/%m/%Y")) {
    miss <- is.na(result) & !is.na(x)
    if (!any(miss)) break
    result[miss] <- suppressWarnings(as.Date(x[miss], format = fmt))
  }
  result
}

# Date outliers are retained in the raw data but excluded from analytical date fields.
# This prevents sentinel values such as 2099-01-01 and malformed historical dates
# from distorting runway calculations and plots.
clean_date_outliers <- function(x, min_date = date_min, max_date = date_max) {
  x <- as.Date(x)
  x[x < min_date | x > max_date] <- as.Date(NA)
  x
}


parse_numeric_safe <- function(x)
  suppressWarnings(as.numeric(str_replace_all(as.character(x), "[$,]", "")))


research_pattern <- paste(c(
  "investigator initiated","investigator-initiated","research project grant",
  "clinical trial","research grant","\\br01\\b","\\br03\\b","\\br21\\b",
  "\\br34\\b","\\bu01\\b","\\bp01\\b","\\bp20\\b","\\bp30\\b","\\bp50\\b",
  "\\bk01\\b","\\bk08\\b","\\bk23\\b","\\bt32\\b","\\bf31\\b","\\bf32\\b"),
  collapse="|")


restricted_pattern <- regex(paste(c(
  "only the following applicant","only the following entit",
  "single source","sole source","invited applicant","not a competitive"),
  collapse="|"), ignore_case=TRUE)


date_cols <- intersect(
  c("post_date", "close_date", "archive_date", "forecasted_post_date",
    "forecasted_close_date", "forecasted_award_date",
    "forecasted_project_start_date"),
  names(grants_raw)
)

num_cols <- intersect(
  c("award_floor", "award_ceiling", "estimated_total_program_funding",
    "expected_number_of_awards"),
  names(grants_raw)
)

# Numeric fields may also be absent from some Grants.gov exports.
for (nm in setdiff(
  c("award_floor", "award_ceiling",
    "estimated_total_program_funding", "expected_number_of_awards"),
  names(grants_raw)
)) {
  grants_raw[[nm]] <- NA_real_
}


grants_parsed <- grants_raw |>
  mutate(across(all_of(date_cols), parse_date_safe),
         across(all_of(num_cols), parse_numeric_safe))

if (length(date_cols) > 0) {
  date_outlier_matrix <- sapply(
    date_cols,
    function(nm) !is.na(grants_parsed[[nm]]) &
      (grants_parsed[[nm]] < date_min | grants_parsed[[nm]] > date_max)
  )
  date_outlier_flag <- rowSums(as.matrix(date_outlier_matrix)) > 0
} else {
  date_outlier_flag <- rep(FALSE, nrow(grants_parsed))
}

grants <- grants_parsed |>
  mutate(
    date_outlier_flag = date_outlier_flag,
    across(all_of(date_cols), clean_date_outliers),
    is_forecast = safe_false(is_forecast),
    across(any_of(c("opportunity_title","agency_name","top_level_agency_name",
                    "summary_description","applicant_types","funding_instruments",
                    "funding_categories","applicant_eligibility_description")),
           ~ str_squish(as.character(.x))),


    status_group = case_when(
      is_forecast                        ~ "Forecasted",
      str_to_lower(coalesce(opportunity_status, "")) == "forecasted" ~ "Forecasted",
      TRUE                               ~ "Posted"),


    agency_group = case_when(
      str_to_lower(str_squish(top_level_agency_name)) ==
        "department of health and human services" ~ "DHHS",
      str_to_lower(str_squish(top_level_agency_name)) ==
        "u.s. national science foundation" ~ "NSF",
      str_to_lower(str_squish(top_level_agency_name)) ==
        "department of defense" ~ "DOD",
      str_to_lower(str_squish(top_level_agency_name)) ==
        "department of agriculture" ~ "USDA",
      str_to_lower(str_squish(top_level_agency_name)) ==
        "department of labor" ~ "DOL",
      str_to_lower(str_squish(top_level_agency_name)) ==
        "department of homeland security" ~ "DHS",
      str_to_lower(str_squish(top_level_agency_name)) ==
        "department of transportation" ~ "DOT",
      TRUE ~ coalesce(top_level_agency_name, "Unknown")
    ),


    applicant_types_text = str_to_lower(coalesce(applicant_types,"")),
    funding_text = str_to_lower(str_c(
      coalesce(opportunity_title,""), coalesce(summary_description,""),
      coalesce(applicant_eligibility_description,""),
      coalesce(category,""), coalesce(category_explanation,""), sep=" | ")),


    state_eligible = str_detect(applicant_types_text, fixed("state_governments")),
    is_nih = str_detect(str_to_lower(str_c(
      coalesce(agency_name,""), coalesce(top_level_agency_name,""),
      coalesce(opportunity_title,""), sep=" | ")),
      "\\bnih\\b|national institutes of health|hhs-nih"),
    is_research_mechanism     = str_detect(funding_text, research_pattern),
    is_restricted_competition = str_detect(funding_text, restricted_pattern),


    days_to_close = as.numeric(close_date - ref_date),
    date_order_valid = is.na(post_date) | is.na(close_date) | close_date >= post_date,
    has_sufficient_runway = is.na(close_date) | days_to_close >= min_runway_days,
    closes_soon = !is.na(days_to_close) & days_to_close >= 0 & days_to_close <= 30,


    close_window = case_when(
      is.na(close_date)    ~ "No stated deadline",
      days_to_close < 0    ~ "Already closed",
      days_to_close <= 30  ~ "0-30 days",
      days_to_close <= 60  ~ "31-60 days",
      days_to_close <= 120 ~ "61-120 days",
      TRUE                 ~ "More than 120 days"),


    award_size = coalesce(award_ceiling, award_floor),
    award_size_group = case_when(
      is.na(award_size)     ~ "Not reported",
      award_size < 250000   ~ "Under $250K",
      award_size < 1000000  ~ "250K-1M",
      award_size < 5000000  ~ "1M-5M",
      award_size < 25000000 ~ "5M-25M",
      TRUE                  ~ "$25M+")
  )


# Determine if the column exists
cost_col <- intersect("cost_sharing_or_matching_requirement", names(grants))

if (length(cost_col) > 0) {
  grants <- grants |> 
    mutate(is_cost_sharing = str_detect(
      str_to_lower(coalesce(.data[[cost_col]], "")), "yes|true|required"
    ))
} else {
  grants <- grants |> 
    mutate(is_cost_sharing = FALSE)
}
```

```{r date-outlier-audit}
date_outlier_audit <- tibble(
  measure = c(
    "Posted rows with close date before reference date",
    "Posted rows with 0-400 days remaining",
    "Posted rows with more than 400 days remaining",
    "Posted rows with no close date",
    "Rows with close date before post date",
    "Rows with invalid date order or out-of-range analytical dates"
  ),
  n = c(
    sum(
      grants$status_group == "Posted" &
        !is.na(grants$close_date) &
        grants$close_date < ref_date,
      na.rm = TRUE
    ),

    sum(
      grants$status_group == "Posted" &
        !is.na(grants$days_to_close) &
        dplyr::between(
          grants$days_to_close,
          0,
          plot_max_runway_days
        ),
      na.rm = TRUE
    ),

    sum(
      grants$status_group == "Posted" &
        !is.na(grants$days_to_close) &
        grants$days_to_close > plot_max_runway_days,
      na.rm = TRUE
    ),

    sum(
      grants$status_group == "Posted" &
        is.na(grants$close_date),
      na.rm = TRUE
    ),

    sum(
      !is.na(grants$post_date) &
        !is.na(grants$close_date) &
        grants$close_date < grants$post_date,
      na.rm = TRUE
    ),

    sum(
      safe_false(grants$date_outlier_flag) |
        !safe_false(grants$date_order_valid),
      na.rm = TRUE
    )
  )
)

date_outlier_audit |>
  DT::datatable(
    rownames = FALSE,
    options = list(
      dom = "t",
      pageLength = 10
    ),
    class = "cell-border stripe hover"
  )
```

## Data quality


```{r data-quality}
tibble(
  measure = c("Rows","Columns","Posted","Forecasted",
              "With close date","With award ceiling",
              "With applicant eligibility","With summary"),
  value   = c(nrow(grants), ncol(grants),
              sum(grants$status_group=="Posted",    na.rm=TRUE),
              sum(grants$status_group=="Forecasted",na.rm=TRUE),
              sum(!is.na(grants$close_date)),
              sum(!is.na(grants$award_ceiling)),
              sum(!is.na(grants$applicant_eligibility_description)),
              sum(!is.na(grants$summary_description)))
) |>
  DT::datatable(rownames=FALSE, options=list(dom="t", pageLength=10))
```


# 1. Executive Summary


```{r exec-summary, results='asis'}
candidate_data <- grants |>
  filter(
    status_group == "Posted",
    if (is_unset(target_agencies)) TRUE
    else agency_group %in% target_agencies,
    state_eligible,
    !is_research_mechanism,
    !is_restricted_competition,
    is.na(close_date) | close_date >= earliest_close,
    is.na(award_size) | award_size >= award_floor_usd
  )

candidate_count <- nrow(candidate_data)


cat('<div class="kpi-grid">',
  '<div class="kpi-card"><div class="kpi-label">Total opportunities</div>',
  '<div class="kpi-value">',comma(nrow(grants)),'</div>',
  '<div class="kpi-note">Posted and forecasted</div></div>',
  '<div class="kpi-card"><div class="kpi-label">Posted</div>',
  '<div class="kpi-value">',comma(sum(grants$status_group=="Posted",na.rm=TRUE)),'</div>',
  '<div class="kpi-note">In snapshot</div></div>',
  '<div class="kpi-card success"><div class="kpi-label">State-eligible</div>',
  '<div class="kpi-value">',comma(sum(grants$state_eligible,na.rm=TRUE)),'</div>',
  '<div class="kpi-note">Initial screen only</div></div>',
  '<div class="kpi-card danger"><div class="kpi-label">NIH tagged</div>',
  '<div class="kpi-value">',comma(sum(grants$is_nih,na.rm=TRUE)),'</div>',
  '<div class="kpi-note">Research-noise source</div></div>',
  '<div class="kpi-card warning"><div class="kpi-label">Screened candidates</div>',
  '<div class="kpi-value">',comma(candidate_count),'</div>',
  '<div class="kpi-note">After all filters</div></div>',
  '</div>')
```


<div class="callout">
This dashboard is an initial screening view. Eligibility tags are broad; the
research-mechanism screen is text-based and must be validated on a random sample
before excluding any opportunity from a final shortlist.
</div>


# 2. Screening Funnel


Each stage is applied only to rows that passed every previous stage.


```{r funnel-prep}
agency_pass <- if (is_unset(target_agencies)) {
  rep(TRUE, nrow(grants))
} else {
  grants$agency_group %in% target_agencies |
    grants$top_level_agency_name %in% target_agencies
}
status_pass      <- if (keep_forecasts) rep(TRUE,nrow(grants)) else grants$status_group=="Posted"
eligibility_pass <- safe_false(grants$state_eligible)
research_pass    <- if (isTRUE(exclude_research_mechanisms))
                      !safe_false(grants$is_research_mechanism) else rep(TRUE,nrow(grants))
restricted_pass  <- if (isTRUE(exclude_restricted_competitions))
                      !safe_false(grants$is_restricted_competition) else rep(TRUE,nrow(grants))
runway_pass <- if (is.null(min_runway_days)) {
  rep(TRUE, nrow(grants))
} else {
  is.na(grants$close_date) |
    (!is.na(grants$days_to_close) &
       grants$days_to_close >= min_runway_days)
}
award_pass       <- if (is.null(award_floor_usd)) rep(TRUE,nrow(grants)) else
                    is.na(grants$award_size) | grants$award_size >= award_floor_usd


funnel_rules <- list(
  `All opportunities`      = rep(TRUE,nrow(grants)),
  `Selected agency`        = agency_pass,
  `Posted`                 = status_pass,
  `State eligible`         = eligibility_pass,
  `Not research mechanism` = research_pass,
  `Open competition`       = restricted_pass,
  `Enough runway`          = runway_pass,
  `Above award threshold`  = award_pass)


survivors <- rep(TRUE, nrow(grants))
funnel_data <- purrr::imap_dfr(funnel_rules, function(rule, name) {
  survivors <<- survivors & rule
  tibble(stage=name, n=sum(survivors, na.rm=TRUE))
}) |>
  mutate(
    dropped = c(0, head(n,-1) - tail(n,-1)),
    pct_all = n / first(n),
    stage   = factor(stage, levels=rev(stage)),
    label   = paste0(comma(n)," (",percent(pct_all, accuracy=0.1),")")
  )
funnel_data
```


```{r funnel-chart, fig.width=11, fig.height=7}
ggplot(funnel_data, aes(x=n, y=stage)) +
  geom_col(fill=pal$blue, width=.68) +
  geom_text(aes(label=label), hjust=-.08, size=4, color=pal$dark) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.18))) +
  labs(title="Sequential Screening Funnel",
       subtitle=paste("Reference date:", format(ref_date,"%B %d, %Y"),
                      "| Minimum runway:", min_runway_days, "days"),
       x="Opportunities remaining", y=NULL) +
  theme(axis.text.y=element_text(face="bold"), panel.grid.major.y=element_blank())
```


```{r funnel-table}
funnel_data |>
  mutate(across(c(n,dropped), comma),
         pct_all=percent(pct_all, accuracy=0.1),
         stage=as.character(stage)) |>
  select(stage, n, dropped, pct_all) |>
  DT::datatable(rownames=FALSE,
    colnames=c("Stage","Remaining","Removed at stage","% of original"),
    options=list(dom="t", pageLength=10, ordering=FALSE),
    class="cell-border stripe hover")
```


# 3. Opportunity Volume by Agency


```{r agency-vol, fig.width=11, fig.height=8}
agency_dat <- grants |>
  count(agency_group, status_group, name="n") |>
  group_by(agency_group) |> mutate(total=sum(n)) |> ungroup() |>
  filter(total >= 5) |>
  mutate(agency_group=fct_reorder(agency_group, total))


ggplot(agency_dat, aes(x=n, y=agency_group, fill=status_group)) +
  geom_col(position="stack", width=.72) +
  geom_text(data=agency_dat |> group_by(agency_group) |>
              summarise(total=first(total),.groups="drop"),
            aes(x=total, y=agency_group, label=comma(total)),
            inherit.aes=FALSE, hjust=-.12, size=3.5, color=pal$dark) +
  scale_fill_manual(values=c(Posted=pal$blue,Forecasted=pal$purple)) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.14))) +
  labs(title="Federal Opportunity Volume by Agency",
       subtitle="Agencies with < 5 opportunities omitted",
       x="Opportunities", y=NULL, fill=NULL) +
  theme(legend.position="top", panel.grid.major.y=element_blank())
```


# 4. Application Deadline Urgency


```{r deadline-urgency-data}
deadline_dat <- grants |>
  filter(
    status_group == "Posted",
    !is.na(close_date),
    !is.na(days_to_close),
    days_to_close >= 0
  ) |>
  mutate(
    close_window = factor(
      case_when(
        days_to_close <= 30 ~ "0–30 days",
        days_to_close <= 60 ~ "31–60 days",
        days_to_close <= 120 ~ "61–120 days",
        days_to_close <= plot_max_runway_days ~
          "121–400 days",
        TRUE ~
          "More than 400 days"
      ),
      levels = c(
        "0–30 days",
        "31–60 days",
        "61–120 days",
        "121–400 days",
        "More than 400 days"
      )
    )
  ) |>
  count(close_window)
```


```{r deadline-urgency-plot, fig.width=10, fig.height=6}
ggplot(
  deadline_dat,
  aes(
    x = close_window,
    y = n,
    fill = close_window
  )
) +
  geom_col(
    width = 0.7,
    show.legend = FALSE
  ) +
  geom_text(
    aes(label = comma(n)),
    vjust = -0.35,
    size = 4
  ) +
  scale_fill_manual(
    values = c(
      "0–30 days" = pal$red,
      "31–60 days" = pal$orange,
      "61–120 days" = pal$blue,
      "121–400 days" = pal$green,
      "More than 400 days" = pal$gray
    )
  ) +
  scale_y_continuous(
    labels = comma,
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Posted Opportunities by Remaining Application Window",
    subtitle = paste(
      "From",
      format(ref_date, "%B %d, %Y"),
      "| Dates are shown through",
      plot_max_runway_days,
      "days"
    ),
    x = NULL,
    y = "Opportunities"
  ) +
  theme(
    axis.text.x = element_text(
      angle = 20,
      hjust = 1
    ),
    panel.grid.major.x = element_blank()
  )

```


# 5. Award Amount Distribution


Award values span orders of magnitude; the x-axis is logarithmic. A national program
ceiling is **not** the amount North Carolina would receive — it is a total across all
awardees nationwide.


```{r award-dist, fig.width=11, fig.height=6}
award_dist_dat <- grants |> filter(status_group=="Posted", !is.na(award_size), award_size > 0)


ggplot(award_dist_dat, aes(x=award_size)) +
  geom_histogram(bins=35, fill=pal$blue, color="white", alpha=.9) +
  geom_vline(xintercept=award_floor_usd, color=pal$red, linetype="dashed", linewidth=1) +
  annotate("text", x=award_floor_usd, y=Inf,
           label=paste0("Threshold: ",dollar(award_floor_usd)),
           hjust=-.05, vjust=1.5, color=pal$red, size=3.5) +
  scale_x_log10(labels=dollar_format(scale_cut=cut_short_scale())) +
  scale_y_continuous(labels=comma) +
  labs(title="Distribution of Reported Award Amounts (Posted)",
       subtitle="Uses award ceiling when available; award floor otherwise",
       x="Award amount (log scale)", y="Opportunities") +
  theme(panel.grid.minor.x=element_blank())
```


# 6. Funding Instruments


A **cooperative agreement** carries ongoing federal involvement and is an early
signal of higher administrative burden compared to a standard grant.


```{r instruments, fig.width=10, fig.height=6}
grants |>
  filter(status_group=="Posted") |>
  mutate(instrument = case_when(
    str_detect(str_to_lower(coalesce(funding_instruments,"")), "cooperative") ~ "Cooperative agreement",
    str_detect(str_to_lower(coalesce(funding_instruments,"")), "\\bgrant\\b")  ~ "Grant",
    str_detect(str_to_lower(coalesce(funding_instruments,"")), "procurement|contract") ~ "Procurement contract",
    str_detect(str_to_lower(coalesce(funding_instruments,"")), "other") ~ "Other",
    TRUE ~ "Not reported")) |>
  count(instrument, sort=TRUE) |>
  mutate(instrument=fct_reorder(instrument,n)) |>
  ggplot(aes(x=n, y=instrument, fill=instrument)) +
  geom_col(width=.7, show.legend=FALSE) +
  geom_text(aes(label=comma(n)), hjust=-.12, size=4) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.14))) +
  labs(title="Posted Opportunities by Funding Instrument",
       subtitle="Cooperative agreements imply ongoing federal oversight",
       x="Posted opportunities", y=NULL) +
  theme(panel.grid.major.y=element_blank())
```


# 7. State Eligibility and Research Noise


<div class="callout callout-green">
**What does "state_governments" eligibility mean?**
Grants.gov uses a controlled vocabulary for applicant types. The code
`state_governments` indicates that the posting explicitly lists state government
entities as eligible applicants. This is one of approximately 15 applicant-type
codes in the schema. **It does not mean the state is the sole or even primary
intended recipient.** Most postings also include universities, hospitals,
nonprofits, tribes, and for-profits — making eligibility alone a very weak
discriminator. The research-mechanism screen below is needed to separate
administrative grants from investigator-initiated research.
</div>


```{r eligibility-research-data}
elig_res <- grants |>
  mutate(
    elig_grp     = if_else(safe_false(state_eligible), "State eligible","Not state eligible"),
    research_grp = if_else(safe_false(is_research_mechanism),"Research screen","Not research screen")
  ) |>
  count(elig_grp, research_grp) |>
  group_by(elig_grp) |> mutate(pct=n/sum(n)) |> ungroup()
```


```{r eligibility-research-plot, fig.width=10, fig.height=6}
ggplot(elig_res, aes(x=elig_grp, y=n, fill=research_grp)) +
  geom_col(position="stack", width=.68) +
  geom_text(aes(label=paste0(comma(n),"\n",percent(pct,accuracy=0.1))),
            position=position_stack(vjust=.5), color="white", fontface="bold", size=3.5) +
  scale_fill_manual(values=c(
    "Research screen" = pal$red,
    "Not research screen" = pal$blue
  )) +
  scale_y_continuous(labels=comma) +
  labs(title="State Eligibility Does Not Identify Administrative Fit",
       subtitle="Research mechanism screen is preliminary and text-based only",
       x=NULL, y="Opportunities", fill=NULL) +
  theme(legend.position="top", panel.grid.major.x=element_blank())
```


<div class="callout callout-warning">
Validate the research screen on a random hand-labelled sample of at least 60
rows before using it to exclude opportunities from the final shortlist.
</div>


# 8. Shortlist After Screening


```{r shortlist-data}
shortlist <- grants |>
  mutate(passed = survivors) |>
  filter(passed) |>
  mutate(
    urgency = case_when(
      is.na(days_to_close) ~ "No stated deadline",
      days_to_close <= 30  ~ "Urgent: 0-30 days",
      days_to_close <= 60  ~ "Near-term: 31-60 days",
      days_to_close <= 120 ~ "Medium-term: 61-120 days",
      TRUE                 ~ "Longer runway"),
    award_display = case_when(
      !is.na(award_ceiling) ~ paste0("Ceiling: ",dollar(award_ceiling)),
      !is.na(award_floor)   ~ paste0("Floor: ",  dollar(award_floor)),
      TRUE                  ~ "Not reported"),
    desc_trunc = str_trunc(coalesce(summary_description,"No summary"), 280)
  ) |>
  arrange(days_to_close, desc(award_size)) |>
  select(opportunity_number, opportunity_title, agency_group,
         urgency, close_date, days_to_close, award_display,
         state_eligible, is_nih, is_research_mechanism,
         opportunity_assistance_listings, desc_trunc, additional_info_url)
```


```{r shortlist-table}
shortlist |>
  mutate(close_date=if_else(is.na(close_date),"No deadline",format(close_date,"%Y-%m-%d"))) |>
  DT::datatable(rownames=FALSE, escape=FALSE, filter="top",
    extensions=c("Buttons","FixedHeader"),
    options=list(pageLength=15, scrollX=TRUE, fixedHeader=TRUE,
                 dom="Bfrtip", buttons=c("copy","csv","excel"),
                 searchHighlight=TRUE),
    class="cell-border stripe hover") |>
  DT::formatStyle("urgency", backgroundColor=DT::styleEqual(
    c("Urgent: 0-30 days","Near-term: 31-60 days",
      "Medium-term: 61-120 days","Longer runway","No stated deadline"),
    c("#FDE2E2","#FFF0D5","#E7F0FA","#E6F4EA","#F2F4F7")))
```


# 9. Opportunities Closing Soon


All posted opportunities closing within 30 days — regardless of whether they
pass the full screening funnel.


```{r urgent-table}
grants |>
  filter(status_group=="Posted", !is.na(close_date),
         close_date >= ref_date, days_to_close <= 30) |>
  mutate(
    award_display = case_when(
      !is.na(award_ceiling) ~ dollar(award_ceiling),
      !is.na(award_floor)   ~ dollar(award_floor),
      TRUE ~ "Not reported"),
    desc_trunc = str_trunc(coalesce(summary_description,"No summary"), 220)
  ) |>
  arrange(days_to_close) |>
  select(opportunity_number, opportunity_title, agency_group,
         close_date, days_to_close, state_eligible, is_nih,
         is_research_mechanism, award_display, desc_trunc) |>
  DT::datatable(rownames=FALSE, filter="top",
    extensions=c("Buttons","FixedHeader"),
    options=list(pageLength=15, scrollX=TRUE, fixedHeader=TRUE,
                 dom="Bfrtip", buttons=c("copy","csv","excel")),
    class="cell-border stripe hover")
```


# 10. Close-Date vs Award Size


Outliers above the 95th percentile of award size (mostly large national NIH and
SAMHSA program totals) are removed to keep the scale informative for
state-administrable grants.


```{r runway-award-data}
runway_award_dat <- grants |>
  filter(
    status_group == "Posted",
    !is.na(close_date),
    !is.na(days_to_close),
    days_to_close >= 0,
    days_to_close <= plot_max_runway_days,
    !is.na(award_size),
    award_size > 0
  ) |>
  mutate(
    elig_disp = if_else(
      safe_false(state_eligible),
      "State eligible",
      "Not state eligible"
    ),

    res_disp = if_else(
      safe_false(is_research_mechanism),
      "Research screen",
      "Not research screen"
    )
  )

p95 <- quantile(
  runway_award_dat$award_size,
  plot_award_quantile,
  na.rm = TRUE
)

runway_award_trim <- runway_award_dat |>
  filter(
    award_size <= p95
  )
```


```{r runway-award-plot, fig.width=11, fig.height=7}
ra_plot <- ggplot(runway_award_trim,
  aes(x=days_to_close, y=award_size,
      color=elig_disp, shape=res_disp, text=opportunity_title)) +
  geom_point(alpha=.7, size=2.5) +
  geom_vline(xintercept=min_runway_days,
             linetype="dashed", color=pal$red, linewidth=.9) +
  annotate("text", x=min_runway_days, y=Inf,
           label=paste0(min_runway_days,"-day runway"),
           hjust=1.05, vjust=1.5, color=pal$red, size=3.2) +
  scale_x_continuous(
  labels = comma,
  limits = c(0, plot_max_runway_days),
  breaks = c(0, 30, 60, 120, 180, 270, 365, 400)
  ) +
  scale_y_continuous(labels=dollar_format(scale_cut=cut_short_scale())) +
  scale_color_manual(values=c(
    "State eligible" = pal$green,
    "Not state eligible" = pal$gray
  )) +
  scale_shape_manual(values=c("Research screen"=17,"Not research screen"=16)) +
  labs(title="Remaining Runway vs Award Size (outliers removed)",
       subtitle = paste0(
  "Runway limited to 0–",
  plot_max_runway_days,
  " days; award amounts above the ",
  plot_award_quantile * 100,
  "th percentile were omitted"
)
,
       x="Days until close date", y="Award size", color=NULL, shape=NULL) +
  theme(legend.position="top")


ggplotly(ra_plot, tooltip=c("x","y","colour","shape","text"))
```


# 11. Most Frequent Assistance Listings


Assistance listings (CFDA numbers) are the join key for USAspending award history.
Programs appearing repeatedly in the snapshot are the highest-priority lookup targets.


```{r cfda-plot, fig.width=11, fig.height=9}
grants |>
  filter(!is.na(opportunity_assistance_listings),
         str_squish(opportunity_assistance_listings) != "") |>
  separate_rows(opportunity_assistance_listings, sep="\\s*;\\s*") |>
  mutate(opportunity_assistance_listings=str_squish(opportunity_assistance_listings)) |>
  filter(opportunity_assistance_listings != "") |>
  count(opportunity_assistance_listings, sort=TRUE, name="n") |>
  slice_head(n=25) |>
  mutate(opportunity_assistance_listings=fct_reorder(opportunity_assistance_listings,n)) |>
  ggplot(aes(x=n, y=opportunity_assistance_listings)) +
  geom_col(fill=pal$purple, width=.7) +
  geom_text(aes(label=comma(n)), hjust=-.12, size=3.3) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.15))) +
  labs(title="Most Frequent Assistance Listings (top 25)",
       subtitle="These CFDA numbers are the priority set for USAspending joins",
       x="Opportunities in snapshot", y=NULL) +
  theme(panel.grid.major.y=element_blank())
```


# 12. Preliminary Triage Quadrants


This view plots every posted opportunity with a stated deadline and reported award
amount against two observable axes: **days of runway** and **award size**. It is a
workload and prioritisation diagnostic — not a measure of need, fit, or probability
of winning.


**How to read the quadrants:**


| Quadrant | Description | Suggested action |
|---|---|---|
| Large award, sufficient runway | High potential value, time to prepare | Assign owner; begin fit assessment |
| Large award, short runway | High value but fast-closing | Leadership decision this week: pursue or formally decline |
| Smaller award, sufficient runway | Lower ceiling; adequate time | Route to relevant division for optional review |
| Smaller award, short runway | Limited time and ceiling | Only pursue if strong existing relationship with program |


```{r quadrant-data}
quad_dat <- grants |>
  filter(status_group=="Posted", !is.na(close_date), close_date >= ref_date,
         !is.na(award_size), award_size > 0) |>
  mutate(
    fit = case_when(
      safe_false(state_eligible) & !safe_false(is_research_mechanism) ~ "Initial fit screen",
      safe_false(state_eligible) &  safe_false(is_research_mechanism) ~ "Eligible; research screen",
     !safe_false(state_eligible)                                       ~ "Not state eligible"),
    quadrant = case_when(
      days_to_close <= min_runway_days & award_size >= award_floor_usd ~ "Large award, short runway",
      days_to_close >  min_runway_days & award_size >= award_floor_usd ~ "Large award, sufficient runway",
      days_to_close <= min_runway_days & award_size <  award_floor_usd ~ "Smaller award, short runway",
      TRUE                                                              ~ "Smaller award, sufficient runway")
  )


p95q <- quantile(quad_dat$award_size, .95, na.rm=TRUE)
quad_trim <- quad_dat |> filter(award_size <= p95q)
```


```{r quadrant-plot, fig.width=11, fig.height=7}
qp <- ggplot(quad_trim,
  aes(x=days_to_close, y=award_size,
      color=fit, shape=quadrant, text=opportunity_title)) +
  geom_point(alpha=.7, size=2.5) +
  geom_vline(xintercept=min_runway_days, linetype="dashed", linewidth=.9, color=pal$red) +
  geom_hline(yintercept=award_floor_usd,  linetype="dashed", linewidth=.9, color=pal$orange) +
  annotate("text", x=min_runway_days, y=Inf,
           label=paste0(min_runway_days,"-day runway"),
           hjust=1.05, vjust=1.5, color=pal$red, size=3.0) +
  annotate("text", x=Inf, y=award_floor_usd,
           label=paste0("Floor: ",dollar(award_floor_usd)),
           hjust=1.05, vjust=-.5, color=pal$orange, size=3.0) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(.02,.06))) +
  scale_y_continuous(labels=dollar_format(scale_cut=cut_short_scale())) +
  scale_color_manual(values=c("Initial fit screen"=pal$green,
                               "Eligible; research screen"=pal$red,
                               "Not state eligible"=pal$gray)) +
  labs(title="Preliminary Triage: Runway vs Award Size (outliers removed)",
       subtitle=paste("Reference date:", format(ref_date,"%B %d, %Y"),
                      "| 95th pct cap:", dollar(round(p95q))),
       x="Days until close date", y="Award size",
       color="Initial screen", shape="Quadrant") +
  theme(legend.position="bottom", legend.box="vertical")


ggplotly(qp, tooltip=c("x","y","colour","shape","text"))
```


```{r quadrant-counts}
quad_dat |> count(quadrant, fit, sort=TRUE) |>
  mutate(pct=percent(n/sum(n), accuracy=0.1)) |>
  DT::datatable(rownames=FALSE,
    colnames=c("Quadrant","Screen result","Opportunities","Share"),
    options=list(dom="t", pageLength=15),
    class="cell-border stripe hover")
```


<div class="callout callout-warning">
This chart does not measure need, capacity, or probability of success. It is a
workload-planning tool only. Do not use award ceilings as estimates of what North
Carolina would receive.
</div>


# 13. Forecasted Opportunity Pipeline


Forecasted opportunities are not open for application. They support calendar
planning and allow early internal review of likely requirements.


```{r forecast-pipeline, fig.width=11, fig.height=6}
grants |>
  filter(status_group=="Forecasted") |>
  mutate(
    fd = coalesce(forecasted_post_date, forecasted_close_date),
    fq = if_else(!is.na(fd), paste0(year(fd)," Q",quarter(fd)), "Date missing")
  ) |>
  count(fq) |>
  mutate(fq=factor(fq, levels=unique(fq))) |>
  ggplot(aes(x=fq, y=n)) +
  geom_col(fill=pal$purple, width=.7) +
  geom_text(aes(label=comma(n)), vjust=-.35, size=4) +
  scale_y_continuous(labels=comma, expand=expansion(mult=c(0,.12))) +
  labs(title="Forecasted Opportunity Pipeline by Quarter",
       subtitle="Forecasts support calendar planning before an opportunity posts",
       x="Expected posting quarter", y="Forecasted opportunities") +
  theme(axis.text.x=element_text(angle=35, hjust=1),
        panel.grid.major.x=element_blank())
```


# 14. NCDHHS-Relevant Domain Classification


Each opportunity may match more than one domain. Unmatched rows should be reviewed,
not assumed irrelevant.


```{r domain-patterns}
domain_patterns <- c(
  "Behavioral health and crisis"          = "behavioral health|mental health|suicide|988|crisis|psychiatric|trauma",
  "Opioid and substance use"              = "opioid|substance use|overdose|naloxone|addiction|recovery|syringe",
  "Maternal, infant, reproductive health" = "maternal|infant|pregnan|prenatal|reproductive|family planning|title x",
  "Child and family well-being"           = "child welfare|children|family|foster|early childhood|parenting|child care",
  "Food and nutrition security"           = "food security|nutrition|wic|snap|fns|hunger|food insecurity",
  "Health and direct-care workforce"      = "health workforce|workforce|nursing|direct care|recruitment|retention",
  "Rural access and telehealth"           = "rural|telehealth|telemedicine|underserved|health access",
  "Disaster recovery and resilience"      = "disaster|resilience|recovery|emergency|hurricane|flood|hazard"
)
```


```{r domain-classification}
domain_text_vec <- str_to_lower(str_c(
  coalesce(grants$opportunity_title,""),
  coalesce(grants$summary_description,""),
  coalesce(grants$applicant_eligibility_description,""),
  coalesce(grants$category, ""), coalesce(grants$category_explanation, ""),
  sep=" | "))


domain_flags <- as.data.frame(purrr::map_dfc(
  domain_patterns,
  ~ str_detect(domain_text_vec, regex(.x, ignore_case=TRUE))
))
names(domain_flags) <- names(domain_patterns)


matched_domains_vec <- apply(domain_flags, 1, function(row) {
  m <- names(domain_patterns)[as.logical(row)]
  if (length(m)==0) "No keyword match" else paste(m, collapse="; ")
})


grants_domains <- grants |>
  bind_cols(domain_flags) |>
  mutate(
    n_domain_matches   = rowSums(across(all_of(names(domain_patterns))), na.rm=TRUE),
    matched_domains    = matched_domains_vec,
    domain_match_group = case_when(
      n_domain_matches == 0 ~ "No keyword match",
      n_domain_matches == 1 ~ "One domain match",
      TRUE                  ~ "Multiple domain matches")
  )
```


```{r domain-volume, fig.width=11, fig.height=8}
purrr::map_dfr(names(domain_patterns), function(dn) {
  tibble(domain=dn,
         Posted     = sum(grants_domains$status_group=="Posted"     & safe_false(grants_domains[[dn]])),
         Forecasted = sum(grants_domains$status_group=="Forecasted" & safe_false(grants_domains[[dn]])))
}) |>
  pivot_longer(c(Posted,Forecasted), names_to="status_group", values_to="n") |>
  mutate(domain=fct_reorder(domain, n, .fun=sum)) |>
  ggplot(aes(x=n, y=domain, fill=status_group)) +
  geom_col(position="stack", width=.7) +
  geom_text(data=~group_by(.,domain)|>summarise(total=sum(n),.groups="drop"),
            aes(x=total,y=domain,label=comma(total)),
            inherit.aes=FALSE, hjust=-.12, size=3.4) +
  scale_fill_manual(values=c(Posted=pal$blue,Forecasted=pal$purple)) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.14))) +
  labs(title="Keyword Match Volume by NCDHHS-Relevant Domain",
       subtitle="An opportunity may match more than one domain",
       x="Keyword matches", y=NULL, fill=NULL) +
  theme(legend.position="top", panel.grid.major.y=element_blank())
```


# 22. NCDHHS 2023–2025 Strategic Plan Alignment (Phase 2)


<div class="callout callout-warning">
<strong>Temporal limitation:</strong> The available NCDHHS strategic plan covers 2023–2025,
while the Grants.gov snapshot concerns 2026 opportunities. The plan is therefore used as a
documented historical alignment framework rather than definitive evidence of current 2026
policy. The table below is a <strong>scaffold requiring manual verification</strong> against the
source PDF at <code>data/raw/strategic_plan/ncdhhs-strategic-plan-2023-25.pdf</code>.
All goal, objective, and strategy text must be confirmed by a human reviewer before use.
</div>


<div class="data-source-box">
<strong>Source:</strong> NCDHHS Strategic Plan 2023–2025 (PDF).<br>
<strong>Expected path:</strong> <code>data/raw/strategic_plan/ncdhhs-strategic-plan-2023-25.pdf</code><br>
<strong>Processed output:</strong> <code>data/processed/ncdhhs_priorities.csv</code><br>
<strong>Status:</strong> Scaffold rows generated; human verification required. Set
<code>human_verified = TRUE</code> and add reviewer initials in <code>reviewer_notes</code>
when each row is confirmed against the source PDF.
</div>


```{r strategic-plan-load}
priorities_path <- file.path(data_dir, "processed", "ncdhhs_priorities.csv")


if (!file.exists(priorities_path)) {
  scaffold <- purrr::map_dfr(paste0("G", 1:5), function(gid) {
    purrr::map_dfr(1:2, function(oi) {
      purrr::map_dfr(1:2, function(si) {
        tibble(
          goal_id        = gid,
          goal_text      = "HUMAN VERIFICATION REQUIRED — populate from source PDF",
          objective_id   = paste0(gid, ".", oi),
          objective_text = "HUMAN VERIFICATION REQUIRED",
          strategy_id    = paste0(gid, ".", oi, ".", si),
          strategy_text  = "HUMAN VERIFICATION REQUIRED",
          source_page    = NA_character_,
          source_document = "ncdhhs-strategic-plan-2023-25.pdf",
          human_verified  = "FALSE",
          reviewer_notes  = paste(
            "Scaffold — populate all text and page references from source PDF.",
            "Set human_verified=TRUE with reviewer initials when complete."),
          priority_domain = NA_character_
        )
      })
    })
  })
  readr::write_csv(scaffold, priorities_path)
  message("Wrote scaffold ncdhhs_priorities.csv — manual verification required.")
}


ncdhhs_priorities <- readr::read_csv(priorities_path, show_col_types=FALSE)
priorities_available <- nrow(ncdhhs_priorities) > 0
```


```{r strategic-plan-table, eval=TRUE}
ncdhhs_priorities |>
  mutate(
    human_verified  = if_else(safe_false(as.logical(human_verified)),
                               "✓ Verified", "⚠ Needs review"),
    priority_domain = coalesce(priority_domain, "—")
  ) |>
  select(goal_id, objective_id, strategy_id, goal_text,
         objective_text, strategy_text, source_page,
         human_verified, priority_domain, reviewer_notes) |>
  DT::datatable(
    rownames=FALSE, filter="top",
    extensions=c("Buttons","FixedHeader"),
    options=list(pageLength=10, scrollX=TRUE, fixedHeader=TRUE,
                 dom="Bfrtip", buttons=c("copy","csv","excel")),
    class="cell-border stripe hover"
  ) |>
  DT::formatStyle("human_verified",
    color=DT::styleEqual(c("✓ Verified","⚠ Needs review"), c("green","orange")),
    fontWeight="bold")
```


# 23. Program-Domain Taxonomy Reference (Phase 3)


The eight program domains connect NCDHHS strategic priorities to Grants.gov
opportunity text, county-level need, county-level capacity, and the final shortlist.
The full taxonomy — including inclusion rules, exclusion rules, need indicators,
capacity indicators, division ownership, and known overlaps — is maintained in
`docs/domain_taxonomy.md`.


<div class="data-source-box">
<strong>Taxonomy document:</strong> <code>docs/domain_taxonomy.md</code><br>
<strong>Status:</strong> Draft v1.0 — pending Phase 2 manual verification and Phase 4 keyword validation.<br>
<strong>Key disambiguation rule:</strong> <code>recovery</code> alone → substance-use (D01).
Requires <code>hurricane</code>, <code>flood</code>, <code>disaster</code>, <code>FEMA</code>,
or <code>hazard mitigation</code> to be classified as disaster recovery (D08).
</div>


```{r domain-taxonomy-display}
tibble(
  ID      = paste0("D0", 1:8),
  Domain  = c(
    "Opioid and substance-use response",
    "Behavioral health and crisis services",
    "Maternal, infant, and reproductive health",
    "Child welfare and family well-being",
    "Food and nutrition security",
    "Health and direct-care workforce",
    "Rural access and telehealth",
    "Disaster recovery and resilience"),
  `Primary division` = c(
    "Behavioral Health (DMH/DD/SUS)",
    "Behavioral Health (DMH/DD/SUS)",
    "Public Health — Women's & Children's Health",
    "Social Services (DSS)",
    "Social Services / Public Health",
    "Office of Rural Health / Health Workforce",
    "Office of Rural Health / Cross-division",
    "Emergency Management / Public Health"),
  `Key need indicators` = c(
    "Overdose mortality, treatment-access gaps",
    "Frequent mental distress, MH HPSA count",
    "Maternal mortality, infant mortality, prenatal access",
    "Child poverty, children uninsured, foster burden",
    "Food insecurity rate, poverty, vehicle access",
    "HPSA count, provider density, disability prevalence",
    "Rural %, broadband gaps, vehicle access, poverty",
    "FEMA declarations, housing vulnerability, rural isolation"),
  `Known overlaps` = c(
    "D02, D07", "D01, D07", "D04, D07", "D03, D05",
    "D04, D03", "D02, D07", "D01–D06", "D01, D02, D07")
) |>
  DT::datatable(
    rownames=FALSE,
    options=list(dom="t", pageLength=10, ordering=FALSE, scrollX=TRUE),
    class="cell-border stripe hover"
  )
```


<div class="callout">
Domain keyword rules were validated against a manual review sample in Section 25.
See <strong>Section 26</strong> for agreement rates and documented rule-based errors.
Assign <code>priority_domain</code> values in <code>data/processed/ncdhhs_priorities.csv</code>
after Phase 4 validation is complete.
</div>


# 15. County-Level Need: ACS Indicators


ACS 2019-2023 5-year estimates provide the demographic and socioeconomic baseline
for the need axis. County boundaries come from the Census TIGER/Line 2023 vintage
via the `tigris` package.


<div class="data-source-box">
**Source:** U.S. Census Bureau, American Community Survey 5-year estimates (2023 vintage).
Requires a free Census API key — set `census_api_key` in params or run
`tidycensus::census_api_key("YOUR_KEY")` before rendering.
**Access:** `tidycensus::get_acs()`
**Cached to:** `data/cache/api/acs_nc_county.rds`
</div>


```{r acs-download}
acs_cache         <- file.path(cache_dir, "api", "acs_nc_county.rds")
nc_counties_cache <- file.path(cache_dir, "api", "nc_counties_tigris.rds")


acs_vars <- c(
  total_pop        = "B01003_001",
  poverty_below    = "B17001_002",
  poverty_universe = "B17001_001",
  median_income    = "B19013_001",
  no_internet      = "B28002_013",
  pop_under18      = "B09001_001",
  uninsured_male   = "B27001_006",
  disability_total = "B18101_001"
)


acs_county <- get_cached(
  acs_cache,
  function() {
    if (nchar(params$census_api_key) == 0) {
      message(
        "No Census API key provided — skipping ACS download."
      )
      return(NULL)
    }

    tidycensus::get_acs(
      geography = "county",
      state = "NC",
      variables = acs_vars,
      year = 2023,
      survey = "acs5",
      output = "wide",
      geometry = FALSE,
      progress_bar = FALSE
    ) |>
      janitor::clean_names() |>
      mutate(
        county_name = str_remove(
          name,
          ", North Carolina"
        ),

        county_join = clean_boundary_name(
          county_name
        ),

        poverty_rate = case_when(
          poverty_universe_e > 0 ~
            poverty_below_e / poverty_universe_e,

          TRUE ~ NA_real_
        ),

        med_income = median_income_e,
        no_internet_n = no_internet_e,
        pop = total_pop_e
      ) |>
      filter(
        !is.na(poverty_rate)
      )
  }
)


nc_counties_sf <- nc_county_boundaries

acs_available <- !is.null(acs_county) && nrow(acs_county) > 0
geo_available <- !is.null(nc_counties_sf)
```


```{r acs-bar, fig.width=11, fig.height=7, eval=TRUE}
acs_county |>
  slice_max(poverty_rate, n=20) |>
  mutate(county_name=fct_reorder(county_name, poverty_rate)) |>
  ggplot(aes(x=poverty_rate, y=county_name)) +
  geom_col(fill=pal$red, width=.72) +
  geom_text(aes(label=percent(poverty_rate, accuracy=0.1)), hjust=-.1, size=3.5) +
  geom_vline(xintercept=mean(acs_county$poverty_rate, na.rm=TRUE),
             linetype="dashed", color=pal$dark, linewidth=.8) +
  scale_x_continuous(labels=percent_format(), expand=expansion(mult=c(0,.15))) +
  labs(title="Top 20 NC Counties by Poverty Rate (ACS 2023)",
       subtitle="Dashed line = statewide average. Source: ACS 2019-2023 5-year.",
       x="Population below poverty level", y=NULL) +
  theme(panel.grid.major.y=element_blank())
```


```{r acs-map, fig.width=11, fig.height=7, eval=TRUE}
map_dat <- nc_counties_sf |>
  left_join(acs_county |> select(geoid, poverty_rate, med_income),
            by=c("GEOID"="geoid"))


ggplot(map_dat) +
  geom_sf(aes(fill=poverty_rate), color="white", linewidth=.3) +
  scale_fill_distiller(palette="YlOrRd", direction=1,
                       labels=percent_format(accuracy=1),
                       na.value="gray90", name="Poverty\nrate") +
  labs(title="NC County Poverty Rate (ACS 2023 5-year)",
       subtitle="Source: Census ACS, B17001. Model-derived; county estimates carry substantial uncertainty.",
       caption="Darker = higher poverty rate") +
  theme_void() +
  theme(legend.position="right",
        plot.title=element_text(face="bold", size=14),
        plot.subtitle=element_text(color=pal$gray))
```


```{r acs-unavailable, results='asis', eval=TRUE}
cat('<div class="callout callout-warning">ACS data not yet downloaded.
Set a Census API key in params (<code>census_api_key</code>) and re-render.</div>')
```


# 16. Provider Shortage Areas: HRSA HPSA


HRSA Health Professional Shortage Areas (HPSAs) are the federal government's own
designation of inadequate delivery capacity. They are difficult to dispute in a
grant application narrative.


<div class="data-source-box">
**Source:** HRSA Bureau of Health Workforce — HPSA Find / Data Downloads.
**Access:** Direct CSV from `data.hrsa.gov`
**Disciplines downloaded:** Mental health, primary care
**Cached to:** `data/cache/api/hrsa_mh.rds`, `hrsa_pc.rds`
</div>


```{r hrsa-download}
hrsa_dl <- function(discipline_code, cache_name) {
  get_cached(file.path(cache_dir, "api", cache_name), function() {
    url <- paste0("https://data.hrsa.gov/DataDownload/DD_Files/",
                  "BCD_HPSA_FCT_DET_", discipline_code, ".csv")
    tryCatch(
      readr::read_csv(url, show_col_types=FALSE, progress=FALSE) |>
        janitor::clean_names(),
      error = function(e) { message("HRSA download failed: ", e$message); NULL }
    )
  })
}


# ── Robust process_hrsa: resolves column names dynamically across HRSA schema versions ──
# FIX: replaces hard-coded hpsa_county_equivalent_name reference that caused crash
process_hrsa <- function(df, discipline) {
  if (is.null(df) || nrow(df) == 0) return(NULL)


  find_col <- function(data, patterns) {
    hits <- names(data)[purrr::map_lgl(names(data), function(n)
      any(str_detect(str_to_lower(n), patterns)))]
    if (length(hits) > 0) hits[1] else NA_character_
  }


  county_col <- find_col(df, c("county_equivalent", "hpsa_county", "county_name"))
  if (is.na(county_col)) county_col <- find_col(df, "county")
  status_col <- find_col(df, "hpsa_status")
  state_col  <- find_col(df, c("common_state_name", "state_name", "state_abbr"))
  type_col   <- find_col(df, c("hpsa_type_name", "type_name"))
  rural_col  <- find_col(df, c("rural_status_name", "rural_status"))
  score_col  <- find_col(df, "hpsa_score")
  id_col     <- find_col(df, "hpsa_id")
  desig_col  <- find_col(df, c("designation_date", "hpsa_designation_date"))
  withd_col  <- find_col(df, c("withdrawal_date",  "hpsa_withdrawal_date"))


  df_filt <- df
  if (!is.na(status_col))
    df_filt <- df_filt |> filter(str_to_upper(.data[[status_col]]) == "DESIGNATED")
  if (!is.na(state_col)) {
    state_values <- str_to_upper(str_squish(as.character(df_filt[[state_col]])))
    df_filt <- df_filt |>
      filter(state_values %in% c("NORTH CAROLINA", "NC"))
  }
  if (nrow(df_filt) == 0) return(NULL)


  df_filt |>
    mutate(
      discipline       = discipline,
      hpsa_score       = if (!is.na(score_col))
                           suppressWarnings(as.numeric(.data[[score_col]])) else NA_real_,
      county_name      = if (!is.na(county_col))
                           str_remove(coalesce(.data[[county_col]], ""),
                                      regex(" County|county", ignore_case=TRUE))
                         else NA_character_,
      shortage_type    = if (!is.na(type_col))  coalesce(.data[[type_col]], "Unknown") else "Unknown",
      rural_status     = if (!is.na(rural_col)) coalesce(.data[[rural_col]], "Unknown") else "Unknown",
      hpsa_id          = if (!is.na(id_col))    as.character(.data[[id_col]])           else NA_character_,
      designation_date = if (!is.na(desig_col)) as.character(.data[[desig_col]])        else NA_character_,
      withdrawal_date  = if (!is.na(withd_col)) as.character(.data[[withd_col]])        else NA_character_
    ) |>
    select(discipline, hpsa_id, county_name, shortage_type,
           rural_status, hpsa_score, designation_date, withdrawal_date)
}


hrsa_mh <- hrsa_dl("MH", "hrsa_mh.rds")
hrsa_pc <- hrsa_dl("PC", "hrsa_pc.rds")


hrsa_nc <- bind_rows(
  process_hrsa(hrsa_mh, "Mental health"),
  process_hrsa(hrsa_pc, "Primary care")
)
hrsa_available <- !is.null(hrsa_nc) && nrow(hrsa_nc) > 0
```


```{r hrsa-county-bar, fig.width=11, fig.height=7, eval=TRUE}
hrsa_county_summary <- hrsa_nc |>
  group_by(county_name, discipline) |>
  summarise(
    n_hpsa = n(),
    max_score = if (all(is.na(hpsa_score))) NA_real_
                else max(hpsa_score, na.rm = TRUE),
    .groups = "drop"
  )


hrsa_county_summary |>
  group_by(county_name) |>
  mutate(total=sum(n_hpsa)) |> ungroup() |>
  slice_max(total, n=40, with_ties=FALSE) |>
  mutate(county_name=fct_reorder(county_name, total)) |>
  ggplot(aes(x=n_hpsa, y=county_name, fill=discipline)) +
  geom_col(position="stack", width=.72) +
  scale_fill_manual(values=c("Mental health"=pal$purple,"Primary care"=pal$blue)) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.12))) +
  labs(title="HRSA Shortage Area Designations by NC County (top 40)",
       subtitle="Each row is one HPSA designation; a county may have several",
       x="HPSA designations", y=NULL, fill="Discipline") +
  theme(panel.grid.major.y=element_blank(), legend.position="top")
```


```{r hrsa-map, fig.width=11, fig.height=7, eval=TRUE}
hrsa_map_dat <- hrsa_nc |>
  filter(discipline=="Mental health") |>
  count(county_name, name="mh_hpsa_count") |>
  mutate(county_name=str_squish(county_name))


nc_map_hrsa <- nc_counties_sf |>
  mutate(county_clean=str_remove(NAME," County| county")) |>
  left_join(hrsa_map_dat, by=c("county_clean"="county_name"))


ggplot(nc_map_hrsa) +
  geom_sf(aes(fill=mh_hpsa_count), color="white", linewidth=.3) +
  scale_fill_distiller(palette="Purples", direction=1,
                       na.value="gray90", name="MH HPSA\ndesignations") +
  labs(title="Mental Health HPSA Designations by NC County",
       subtitle="Source: HRSA HPSA data. Gray = no current designation or county boundary mismatch.") +
  theme_void() +
  theme(plot.title=element_text(face="bold",size=14),
        plot.subtitle=element_text(color=pal$gray))
```


```{r hrsa-table, eval=TRUE}
hrsa_nc |>
  arrange(discipline, desc(hpsa_score)) |>
  DT::datatable(rownames=FALSE, filter="top",
    options=list(pageLength=15, scrollX=TRUE, dom="Bfrtip", buttons=c("copy","csv")),
    class="cell-border stripe hover")
```


```{r hrsa-unavailable, results='asis', eval=TRUE}
cat('<div class="callout callout-warning">HRSA data could not be downloaded.
Check network access to <code>data.hrsa.gov</code> and re-render.</div>')
```


# 17. Local Health Outcomes: CDC PLACES


CDC PLACES provides model-based estimates of health outcomes at county, place,
tract, and ZCTA level. These are **modelled estimates**, not direct survey counts.


<div class="data-source-box">
**Source:** CDC PLACES 2025 release — County GIS-Friendly Format.
**Access:** CDC Open Data Portal (Socrata API), dataset ID `i46a-9kgh`
**Key measures used:** Frequent mental distress, depression prevalence
**Cached to:** `data/cache/api/places_nc_county.rds`
</div>


```{r places-download}
places_cache <- file.path(cache_dir, "api", "places_nc_county.rds")


places_nc <- get_cached(places_cache, function() {
  base_url <- "https://data.cdc.gov/resource/i46a-9kgh.csv"
  query    <- "?StateDesc=North+Carolina&$limit=200"
  tryCatch(
    readr::read_csv(paste0(base_url, query), show_col_types=FALSE, progress=FALSE) |>
      janitor::clean_names(),
    error=function(e) { message("PLACES download failed: ", e$message); NULL }
  )
})


places_available <- !is.null(places_nc) && nrow(places_nc) > 0


if (places_available) {
  key_measures <- c("mhlth", "depression", "binge", "csmoking", "access2", "bphigh")

  # PLACES column names can vary between releases/API exports. Normalize the
  # county identifier first instead of assuming that clean_names() produced
  # exactly `county_name`.
  county_candidates <- c(
    "county_name", "countyname", "county", "locationname", "location_name",
    "name"
  )
  county_col <- county_candidates[county_candidates %in% names(places_nc)][1]

  crude_cols <- names(places_nc)[
    str_detect(
      names(places_nc),
      paste0("^(", paste(key_measures, collapse="|"), ")_(crudeprev|crude_prev)$")
    )
  ]

  if (length(county_col) == 1 && length(crude_cols) > 0) {
    places_nc <- places_nc |>
      rename(county_name = all_of(county_col))

    places_long <- places_nc |>
      select(county_name, any_of(crude_cols)) |>
      pivot_longer(
        cols = -county_name,
        names_to = "measure",
        values_to = "value"
      ) |>
      mutate(value=suppressWarnings(as.numeric(as.character(value)))) |>
      filter(!is.na(value))
  } else {
    message("PLACES county or crude prevalence columns not found; skipping PLACES reshaping.")
    places_available <- FALSE
    places_long <- tibble()
  }
}
```


```{r places-mental-bar, fig.width=11, fig.height=7, eval=TRUE}
mhlth_col <- names(places_nc)[
  str_detect(str_to_lower(names(places_nc)), "mhlth") &
  str_detect(str_to_lower(names(places_nc)), "crude")
][1]


if (length(mhlth_col) == 1 && !is.na(mhlth_col)) {
  places_nc |>
    rename(mhlth_val=all_of(mhlth_col)) |>
    mutate(mhlth_val=suppressWarnings(as.numeric(as.character(mhlth_val)))) |>
    filter(!is.na(mhlth_val)) |>
    slice_max(mhlth_val, n=20) |>
    mutate(county_name=fct_reorder(county_name, mhlth_val)) |>
    ggplot(aes(x=mhlth_val, y=county_name)) +
    geom_col(fill=pal$purple, width=.72) +
    geom_text(aes(label=paste0(round(mhlth_val,1),"%")), hjust=-.1, size=3.5) +
    scale_x_continuous(expand=expansion(mult=c(0,.15))) +
    labs(title="Top 20 NC Counties: Frequent Mental Distress Prevalence",
         subtitle="CDC PLACES 2025 model-based estimate. These are smoothed estimates, not direct observations.",
         x="% adults reporting frequent mental distress", y=NULL) +
    theme(panel.grid.major.y=element_blank())
} else {
  message("Mental distress column not found in PLACES data — check column names.")
}
```


```{r places-map, fig.width=11, fig.height=7, eval=TRUE}
places_county_col <- names(places_nc)[
  str_detect(str_to_lower(names(places_nc)), "county|locationname|location_name")
][1]

places_map <- if (
  length(mhlth_col) == 1 && !is.na(mhlth_col) &&
  length(places_county_col) == 1 && !is.na(places_county_col)
) {
  places_join <- places_nc |>
    transmute(
      locationname = as.character(.data[[places_county_col]]),
      mhlth_val = suppressWarnings(as.numeric(as.character(.data[[mhlth_col]])))
    ) |>
    mutate(locationname = str_remove(locationname, ",?\\s*North Carolina$"))

  nc_counties_sf |>
    mutate(county_clean = str_remove(NAME, " County| county")) |>
    left_join(places_join, by = c("county_clean" = "locationname"))
} else {
  NULL
}

if (!is.null(places_map)) {
  ggplot(places_map) +
    geom_sf(aes(fill = mhlth_val), color = "white", linewidth = .3) +
    scale_fill_distiller(
      palette = "Purples", direction = 1,
      na.value = "gray90", name = "% frequent\nmental distress"
    ) +
    labs(
      title = "Frequent Mental Distress Prevalence by NC County",
      subtitle = "Source: CDC PLACES 2025. Model-based estimate."
    ) +
    theme_void() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      plot.subtitle = element_text(color = pal$gray)
    )
}
```


```{r places-unavailable, results='asis', eval=TRUE}
cat('<div class="callout callout-warning">CDC PLACES data could not be downloaded.
Check network access to <code>data.cdc.gov</code> and re-render.</div>')
```


# 18. Disaster Exposure: FEMA Declarations


Federal disaster declarations identify counties with accumulated disaster exposure.
Hurricane Helene (2024) and other recent events directly affect the disaster-recovery
program domain.


<div class="data-source-box">
**Source:** FEMA OpenFEMA Disaster Declarations Summaries v2.
**Access:** `https://www.fema.gov/api/open/v2/disasterDeclarationsSummaries`
**Scope:** North Carolina, declarations since 2000
**Cached to:** `data/cache/api/fema_nc.rds`
</div>


```{r fema-download}
fema_cache <- file.path(cache_dir, "api", "fema_nc.rds")


fema_nc <- get_cached(fema_cache, function() {
  url <- paste0(
    "https://www.fema.gov/api/open/v2/disasterDeclarationsSummaries",
    "?$filter=state%20eq%20'NC'%20and%20declarationDate%20ge%20'2000-01-01'",
    "&orderby=declarationDate%20desc&format=json&$top=2000"
  )
  resp <- tryCatch(
    httr2::request(url) |>
      httr2::req_timeout(30) |>
      httr2::req_error(is_error=function(r) FALSE) |>
      httr2::req_perform(),
    error=function(e) { message("FEMA request failed: ", e$message); NULL }
  )
  if (is.null(resp) || httr2::resp_status(resp) != 200) return(NULL)
  parsed <- httr2::resp_body_json(resp, simplifyVector=TRUE)
  if (is.null(parsed[["DisasterDeclarationsSummaries"]])) return(NULL)
  as_tibble(parsed[["DisasterDeclarationsSummaries"]]) |>
    janitor::clean_names() |>
    mutate(declaration_date=parse_date_safe(declaration_date),
           decl_year=year(declaration_date))
})


fema_available <- !is.null(fema_nc) && nrow(fema_nc) > 0
```


```{r fema-by-year, fig.width=11, fig.height=6, eval=TRUE}
fema_nc |>
  filter(!is.na(decl_year), decl_year >= 2000) |>
  count(decl_year, declaration_type) |>
  ggplot(aes(x=decl_year, y=n, fill=declaration_type)) +
  geom_col(width=.8, position="stack") +
  scale_fill_manual(values=c(DR=pal$red,EM=pal$orange, FM=pal$blue),na.value=pal$gray) +
  scale_x_continuous(breaks=seq(2000,2030,2)) +
  scale_y_continuous(labels=comma, expand=expansion(mult=c(0,.12))) +
  labs(title="Federal Disaster Declarations in NC by Year",
       subtitle="DR = Major Disaster; EM = Emergency; FM = Fire Management",
       x=NULL, y="Declarations", fill="Type") +
  theme(panel.grid.major.x=element_blank(), legend.position="top",
        axis.text.x=element_text(angle=35, hjust=1))
```


```{r fema-county-map, fig.width=11, fig.height=7, eval=TRUE}
fema_county <- fema_nc |>
  filter(declaration_type=="DR", !is.na(decl_year), decl_year >= 2015,
         !is.na(designated_area)) |>
  mutate(county_clean=str_remove(designated_area," \\(County\\)| County| county")) |>
  count(county_clean, name="dr_count")


nc_fema_map <- nc_counties_sf |>
  left_join(fema_county, by=c("NAME"="county_clean"))


ggplot(nc_fema_map) +
  geom_sf(aes(fill=dr_count), color="white", linewidth=.3) +
  scale_fill_distiller(palette="OrRd", direction=1,
                       na.value="gray90",
                       name="Major disaster\ndeclarations\n(2015-present)") +
  labs(title="NC County Major Disaster Declarations (2015-present)",
       subtitle="Source: FEMA OpenFEMA v2. Includes Hurricane Helene 2024 and prior events.") +
  theme_void() +
  theme(plot.title=element_text(face="bold",size=14),
        plot.subtitle=element_text(color=pal$gray))
```


```{r fema-recent-table, eval=TRUE}
fema_nc |>
  filter(!is.na(decl_year)) |>
  arrange(desc(declaration_date)) |>
  slice_head(n=50) |>
  select(disaster_number, declaration_date, declaration_type,
         incident_type, designated_area, fy_declared) |>
  DT::datatable(rownames=FALSE, filter="top",
    options=list(pageLength=15, scrollX=TRUE, dom="Bfrtip", buttons=c("copy","csv")),
    class="cell-border stripe hover")
```


```{r fema-unavailable, results='asis', eval=TRUE}
cat('<div class="callout callout-warning">FEMA data could not be downloaded.
Check network access to <code>www.fema.gov</code> and re-render.</div>')
```


# 19. Federal Award History: USAspending Peer-State Comparison


For each CFDA program appearing in the grants snapshot, this section queries
USAspending to find which states have received awards in the last five years.
Programs where peer states (GA, VA, SC, TN) received funding but North Carolina
did not are the highest-priority gap candidates.


**Important caveat:** Award history reflects past applications, not eligibility.
A gap may mean NC never applied, applied unsuccessfully, or is ineligible by
practice but not by rule. Treat gaps as conversation starters, not proof of
missed money.


<div class="data-source-box">
**Source:** USAspending.gov API v2 — `spending_by_geography` endpoint.
**Access:** No authentication required.
**Scope:** Top 30 CFDAs by frequency in grants snapshot; FY2021-2025; grants and cooperative agreements.
**Cached to:** `data/cache/api/usa_cfda_*.rds`
</div>


```{r usaspending-download}
top_cfdas <- grants |>
  filter(!is.na(opportunity_assistance_listings),
         str_squish(opportunity_assistance_listings) != "") |>
  separate_rows(opportunity_assistance_listings, sep="\\s*;\\s*") |>
  mutate(cfda=str_extract(str_squish(opportunity_assistance_listings), "^[0-9]+\\.[0-9]+")) |>
  filter(!is.na(cfda)) |>
  count(cfda, sort=TRUE) |>
  slice_head(n=30)


fetch_cfda_states <- function(cfda_num) {
  cache_path <- file.path(cache_dir, "api",
                           paste0("usa_", gsub("\\.","_",cfda_num), ".rds"))
  get_cached(cache_path, function() {
    body <- list(
      scope="place_of_performance", geo_layer="state",
      filters=list(
        award_type_codes=list("02","03","04","05"),
        time_period=list(list(start_date="2021-01-01", end_date="2025-12-31")),
        program_numbers=list(cfda_num)))
    resp <- tryCatch(
      httr2::request("https://api.usaspending.gov/api/v2/search/spending_by_geography/") |>
        httr2::req_body_json(body) |>
        httr2::req_timeout(25) |>
        httr2::req_error(is_error=function(r) FALSE) |>
        httr2::req_perform(),
      error=function(e) NULL)
    if (is.null(resp) || httr2::resp_status(resp) != 200) return(NULL)
    parsed <- httr2::resp_body_json(resp, simplifyVector=TRUE)
    results <- parsed[["results"]]
    if (is.null(results) || length(results) == 0) return(NULL)

    result_tbl <- as_tibble(results)

    required_result_cols <- c(
      "shape_code", "display_name", "aggregated_amount"
    )
    if (!all(required_result_cols %in% names(result_tbl))) return(NULL)

    result_tbl |>
      transmute(
        cfda = cfda_num,
        state_code = as.character(shape_code),
        state_name = as.character(display_name),
        total_awards = suppressWarnings(as.numeric(aggregated_amount))
      )
  })
}


usa_results <- purrr::map(top_cfdas$cfda, function(c) { Sys.sleep(.25); fetch_cfda_states(c) })
usa_df       <- bind_rows(Filter(Negate(is.null), usa_results))
usa_available <- nrow(usa_df) > 0
```


```{r usaspending-gap-analysis, eval=TRUE}
focus_states <- c("NC","GA","VA","SC","TN")


usa_pivot <- usa_df |>
  filter(state_code %in% focus_states) |>
  mutate(received=total_awards > 0) |>
  select(cfda, state_code, received) |>
  pivot_wider(names_from=state_code, values_from=received, values_fill=FALSE)


nc_gaps <- usa_pivot |>
  filter(!coalesce(NC, FALSE)) |>
  rowwise() |>
  mutate(n_peers_received=sum(c_across(any_of(c("GA","VA","SC","TN"))), na.rm=TRUE)) |>
  ungroup() |>
  filter(n_peers_received >= 1) |>
  arrange(desc(n_peers_received)) |>
  left_join(top_cfdas, by="cfda") |>
  rename(snapshot_count=n)
```


```{r usaspending-heatmap, fig.width=11, fig.height=8, eval=TRUE}
heat_dat <- usa_df |>
  filter(state_code %in% focus_states) |>
  mutate(has_awards=if_else(total_awards > 0, "Received", "Not received"),
         cfda=factor(cfda, levels=rev(sort(unique(cfda)))))


ggplot(heat_dat, aes(x=state_code, y=cfda, fill=has_awards)) +
  geom_tile(color="white", linewidth=.4) +
  scale_fill_manual(values=c(
    "Received" = pal$green,
    "Not received" = pal$light
  ), na.value=pal$light) +
  labs(title="Federal Award Receipt by State and CFDA Program (FY 2021-2025)",
       subtitle="Green = at least one award received. Top 30 CFDAs by frequency in snapshot.",
       x="State", y="CFDA number", fill=NULL) +
  theme(panel.grid=element_blank(), legend.position="top", axis.text.y=element_text(size=8))
```


```{r usaspending-gap-table, eval=TRUE}
nc_gaps |>
  select(cfda, n_peers_received, GA, VA, SC, TN, snapshot_count) |>
  mutate(across(c(GA,VA,SC,TN), ~ if_else(.x, "Yes", "No"))) |>
  DT::datatable(rownames=FALSE, filter="top",
    colnames=c("CFDA","Peer states w/ award","GA","VA","SC","TN","In snapshot"),
    options=list(pageLength=15, dom="Bfrtip", buttons=c("copy","csv")),
    class="cell-border stripe hover")
```


```{r usaspending-unavailable, results='asis', eval=TRUE}
cat('<div class="callout callout-warning">USAspending data could not be retrieved.
Check network access to <code>api.usaspending.gov</code> and re-render.</div>')
```


# 20. Behavioral Health Capacity: SAMHSA Facilities


SAMHSA's N-SUMHSS survey provides facility-level counts of substance-use and
mental-health treatment capacity. Missing facilities should not be interpreted
as proof that no services exist — survey participation is voluntary.


<div class="data-source-box">
**Source:** SAMHSA National Survey of Substance Abuse Treatment Services (N-SUMHSS / NSSATS).
**Download:** Visit `samhsa.gov/data` > N-SUMHSS > Download most recent public-use file.
**Expected local path:** `data/raw/samhsa/nsumhss-latest.csv`
**Note:** No unauthenticated public bulk download API exists. File must be downloaded manually.
</div>


```{r samhsa-load}
samhsa_paths <- c(
  file.path(data_dir, "raw", "samhsa", "nsumhss-latest.csv"),
  file.path(data_dir, "raw", "samhsa", "nsumhss-2023.csv"),
  file.path(data_dir, "raw", "samhsa", "nsumhss-2022.csv")
)
samhsa_file      <- Filter(file.exists, samhsa_paths)[1]
samhsa_available <- !is.na(samhsa_file) && length(samhsa_file) > 0


samhsa_nc <- if (samhsa_available) {
  tryCatch({
    dat <- readr::read_csv(samhsa_file, show_col_types = FALSE) |>
      janitor::clean_names()

    state_col <- names(dat)[
      str_detect(str_to_lower(names(dat)), "^state$|state_")
    ][1]

    if (is.na(state_col)) {
      stop("SAMHSA file does not contain a recognizable state column.")
    }

    dat |>
      filter(str_to_upper(str_squish(as.character(.data[[state_col]]))) %in%
               c("NC", "NORTH CAROLINA"))
  }, error = function(e) {
    message("SAMHSA load failed: ", conditionMessage(e))
    samhsa_available <<- FALSE
    NULL
  })
} else NULL
```


```{r samhsa-bar, fig.width=11, fig.height=7, eval=TRUE}
county_col <- names(samhsa_nc)[str_detect(str_to_lower(names(samhsa_nc)), "county")][1]
if (!is.na(county_col)) {
  samhsa_nc |>
    rename(county=all_of(county_col)) |>
    count(county, sort=TRUE) |>
    slice_head(n=20) |>
    mutate(county=fct_reorder(county,n)) |>
    ggplot(aes(x=n, y=county)) +
    geom_col(fill=pal$purple, width=.72) +
    geom_text(aes(label=comma(n)), hjust=-.1, size=3.5) +
    scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.14))) +
    labs(title="SAMHSA Treatment Facilities by NC County (top 20)",
         subtitle="Source: SAMHSA N-SUMHSS. Survey-based; absence does not equal no service.",
         x="Facilities in dataset", y=NULL) +
    theme(panel.grid.major.y=element_blank())
}
```


```{r samhsa-unavailable, results='asis', eval=TRUE}
cat('<div class="callout callout-warning">
SAMHSA N-SUMHSS file not found at any of the expected paths:
<code>data/raw/samhsa/nsumhss-latest.csv</code><br>
Download the most recent public-use file from
<a href="https://www.samhsa.gov/data/data-we-collect/n-sumhss-national-substance-use-and-mental-health-services-survey/datafiles" target="_blank">samhsa.gov/data</a>
and save it to that path, then re-render.
</div>')
```


# 21. Overdose Mortality: CDC WONDER


County-level drug overdose death rates require a manual query from CDC WONDER
because the system does not expose a public REST API. Small-county values are
suppressed by NCHS and must remain missing — do not recode suppressed values as zero.


<div class="data-source-box">
**Source:** CDC WONDER Underlying Cause of Death (ICD-10 codes X40-X44, X60-X64, X85, Y10-Y14).
**Download:** `wonder.cdc.gov` — Underlying Cause of Death 1999-2022, group by county, NC only.
**Expected local path:** `data/raw/cdc_wonder/overdose-mortality-county.txt` or `.csv`
</div>


```{r wonder-load}
wonder_paths <- c(
  file.path(data_dir, "raw", "cdc_wonder", "overdose-mortality-county.txt"),
  file.path(data_dir, "raw", "cdc_wonder", "overdose-mortality-county.csv")
)
wonder_file      <- Filter(file.exists, wonder_paths)[1]
wonder_available <- !is.na(wonder_file) && length(wonder_file) > 0


wonder_nc <- if (wonder_available) {
  tryCatch({
    raw <- if (grepl("\\.csv$", wonder_file[1], ignore.case = TRUE)) {
      readr::read_csv(wonder_file[1], show_col_types = FALSE)
    } else {
      readr::read_tsv(wonder_file[1], show_col_types = FALSE, comment = "---")
    } |>
      janitor::clean_names()
    if (!"county" %in% names(raw) && "county_1" %in% names(raw))
      raw <- rename(raw, county=county_1)
    raw |>
      filter(!is.na(county), county != "Total") |>
      mutate(
        county_name = str_remove(
          str_squish(county),
          ",?\\s*(NC|North Carolina)$"
        ),
        deaths      = suppressWarnings(as.numeric(deaths)),
        crude_rate  = suppressWarnings(as.numeric(crude_rate))
      )
  }, error=function(e) { wonder_available <<- FALSE; NULL })
} else NULL
```


```{r wonder-map, fig.width=11, fig.height=7, eval=TRUE}
wonder_map_dat <- nc_counties_sf |>
  left_join(wonder_nc |> select(county_name, crude_rate), by=c("NAME"="county_name"))


ggplot(wonder_map_dat) +
  geom_sf(aes(fill=crude_rate), color="white", linewidth=.3) +
  scale_fill_distiller(palette="Reds", direction=1,
                       na.value="gray90", name="Crude rate\nper 100k") +
  labs(title="Drug Overdose Mortality Rate by NC County",
       subtitle="Source: CDC WONDER. Gray = suppressed (<10 deaths) or data not matched.") +
  theme_void() +
  theme(plot.title=element_text(face="bold",size=14),
        plot.subtitle=element_text(color=pal$gray))
```


```{r wonder-bar, fig.width=11, fig.height=7, eval=TRUE}
wonder_nc |>
  filter(!is.na(crude_rate)) |>
  slice_max(crude_rate, n=20) |>
  mutate(county_name=fct_reorder(county_name, crude_rate)) |>
  ggplot(aes(x=crude_rate, y=county_name)) +
  geom_col(fill=pal$red, width=.72) +
  geom_text(aes(label=round(crude_rate,1)), hjust=-.1, size=3.5) +
  scale_x_continuous(expand=expansion(mult=c(0,.14))) +
  labs(title="Top 20 NC Counties by Drug Overdose Crude Mortality Rate",
       subtitle="Source: CDC WONDER. Suppressed counties (< 10 deaths) excluded.",
       x="Crude rate per 100,000", y=NULL) +
  theme(panel.grid.major.y=element_blank())
```


```{r wonder-unavailable, results='asis', eval=TRUE}
cat('<div class="callout callout-warning">
CDC WONDER file not found. CDC WONDER does not expose a public API.<br>
<strong>To download:</strong>
<ol>
<li>Go to <a href="https://wonder.cdc.gov" target="_blank">wonder.cdc.gov</a></li>
<li>Select <em>Underlying Cause of Death 1999-2020</em></li>
<li>Group by: State (NC only), County</li>
<li>ICD-10 codes: X40-X44, X60-X64, X85, Y10-Y14</li>
<li>Export as Tab-delimited text and save to
<code>data/raw/cdc_wonder/overdose-mortality-county.txt</code></li>
</ol>
</div>')
```

# 22. Configurable North Carolina Geographic Maps

The map geography is controlled by the `map_geography` parameter at the top of this
file. Supported values are `county`, `congressional_district`, `tract`, and `place`.
Changing that one parameter changes the boundary layer used by the interactive map.

Geographic indicators are never copied from one geography to another. For example,
county poverty estimates remain county estimates; selecting congressional districts
or census tracts changes the displayed boundaries but does not falsely convert the
county estimate into a district- or tract-level estimate.

```{r boundary-data}
boundary_cache_dir <- file.path(cache_dir, "boundaries")
dir.create(boundary_cache_dir, showWarnings = FALSE, recursive = TRUE)

clean_boundary_name <- function(x) {
  x |>
    as.character() |>
    str_remove(regex("\\s+County$", ignore_case = TRUE)) |>
    str_remove(regex(",\\s*North Carolina$", ignore_case = TRUE)) |>
    str_squish() |>
    str_to_upper()
}

get_nc_boundaries <- function(geography, year = boundary_year, resolution = "500k") {
  valid <- c("county", "congressional_district", "tract", "place")
  geography <- tolower(str_squish(geography))
  if (!geography %in% valid) {
    stop("Unsupported map_geography: ", geography,
         ". Choose one of: ", paste(valid, collapse = ", "))
  }

  cache_path <- file.path(
    boundary_cache_dir,
    paste0("nc_", geography, "_", year, ".rds")
  )

  get_cached(cache_path, function() {
    tryCatch({
      result <- switch(
        geography,
        county = tigris::counties(
          state = "NC", cb = TRUE, resolution = resolution,
          year = year, progress_bar = FALSE
        ),
        congressional_district = tigris::congressional_districts(
          state = "NC", cb = TRUE, resolution = resolution,
          year = year, progress_bar = FALSE
        ),
        tract = tigris::tracts(
          state = "NC", cb = TRUE, year = year,
          progress_bar = FALSE
        ),
        place = tigris::places(
          state = "NC", cb = TRUE, year = year,
          progress_bar = FALSE
        )
      )
      sf::st_transform(result, 4326)
    }, error = function(e) {
      message("Boundary download failed for ", geography, ": ", conditionMessage(e))
      NULL
    })
  })
}

selected_boundaries <- get_nc_boundaries(map_geography)

if (!is.null(selected_boundaries) && nrow(selected_boundaries) > 0) {
  district_field <- intersect(c("CD119FP", "CD118FP", "GEOID"), names(selected_boundaries))[1]
  district_values <- if (!is.na(district_field)) {
    as.character(selected_boundaries[[district_field]])
  } else {
    rep(NA_character_, nrow(selected_boundaries))
  }

  selected_boundaries <- selected_boundaries |>
    mutate(
      boundary_label = case_when(
        map_geography == "county" & "NAME" %in% names(selected_boundaries) ~ paste0(NAME, " County"),
        map_geography == "congressional_district" ~ paste0("Congressional District ", district_values),
        map_geography == "tract" & "GEOID" %in% names(selected_boundaries) ~ paste0("Census tract ", GEOID),
        map_geography == "place" & "NAME" %in% names(selected_boundaries) ~ NAME,
        "GEOID" %in% names(selected_boundaries) ~ as.character(GEOID),
        TRUE ~ NA_character_
      ),
      county_join = if ("NAME" %in% names(selected_boundaries)) clean_boundary_name(NAME) else NA_character_,
      map_popup = boundary_label
    )
}

map_boundary_available <- !is.null(selected_boundaries) &&
  nrow(selected_boundaries) > 0

# County poverty is the only currently supplied filled geographic indicator.
# It can be displayed when the selected map geography is county.
if (map_geography == "county" &&
    map_boundary_available &&
    exists("acs_county") &&
    !is.null(acs_county) &&
    nrow(acs_county) > 0) {

  acs_map_data <- selected_boundaries |>
    left_join(
      acs_county |>
        select(any_of(c("county_join", "geoid", "county_name", "poverty_rate", "med_income"))),
      by = "county_join",
      suffix = c("_geometry", "_acs")
    ) |>
    mutate(
      display_name = coalesce(
        .data[[intersect(c("county_name", "county_name_acs"), names(.))[1]]],
        boundary_label
      ),
      poverty_label = if_else(
        is.na(poverty_rate), "Not available", percent(poverty_rate, accuracy = 0.1)
      ),
      income_label = if_else(
        is.na(med_income), "Not available", dollar(med_income)
      ),
      boundary_popup = paste0(
        "<strong>", display_name, "</strong><br/>",
        "Poverty rate: ", poverty_label, "<br/>",
        "Median household income: ", income_label, "<br/><br/>",
        "<em>ACS county estimate.</em>"
      )
    )

  poverty_palette <- leaflet::colorBin(
    palette = "YlOrRd",
    domain = acs_map_data$poverty_rate,
    bins = c(0, 0.10, 0.15, 0.20, 0.25, 1),
    na.color = "#D9D9D9"
  )
} else {
  acs_map_data <- selected_boundaries
  poverty_palette <- NULL
}

if (map_boundary_available) {
  selected_boundaries$map_fill <- "#E8EEF5"
  if (map_geography == "county" &&
      "poverty_rate" %in% names(selected_boundaries) &&
      !is.null(poverty_palette)) {
    selected_boundaries$map_fill <- poverty_palette(selected_boundaries$poverty_rate)
  }
}
```

```{r interactive-boundary-map, eval=TRUE}
map <- leaflet::leaflet(
  data = selected_boundaries,
  options = leafletOptions(minZoom = 5, maxZoom = 12)
) |>
  leaflet::setView(lng = -79.5, lat = 35.5, zoom = 7) |>
  leaflet::addPolygons(
    fillColor = ~map_fill,
    fillOpacity = 0.75,
    color = "#FFFFFF",
    weight = if (map_geography == "tract") 0.5 else 1,
    opacity = 1,
    label = ~boundary_label,
    popup = ~map_popup,
    group = paste0("NC ", map_geography, " boundaries"),
    highlightOptions = leaflet::highlightOptions(
      weight = 2, color = "#2166AC", fillOpacity = 0.9,
      bringToFront = TRUE
    )
  ) |>
  leaflet::addScaleBar(position = "bottomleft")

if (map_geography == "county" && !is.null(poverty_palette)) {
  map <- map |>
    leaflet::addLegend(
      position = "bottomright",
      pal = poverty_palette,
      values = ~poverty_rate,
      title = "County poverty rate",
      opacity = 0.9
    )
}

map
```

```{r boundary-availability}
boundary_availability <- tibble(
  selected_geography = map_geography,
  boundary_year = boundary_year,
  available = map_boundary_available,
  rows = if (map_boundary_available) nrow(selected_boundaries) else 0L
)

boundary_availability |>
  DT::datatable(
    rownames = FALSE,
    options = list(dom = "t", pageLength = 10),
    class = "cell-border stripe hover"
  )
```

## Geographic join diagnostics

These diagnostics are retained for county-level datasets because those datasets can
be joined to county boundaries without changing their geographic meaning.

```{r county-join-diagnostics}
county_join_diagnostics <- tibble(
  dataset = character(),
  source_rows = integer(),
  matched_counties = integer(),
  unmatched_rows = integer(),
  match_rate = numeric()
)

if (exists("acs_county") && !is.null(acs_county) && nrow(acs_county) > 0 &&
    exists("selected_boundaries") && map_geography == "county") {
  acs_join_result <- acs_county |>
    mutate(county_join = clean_boundary_name(county_name))

  matched <- acs_join_result$county_join %in% selected_boundaries$county_join
  county_join_diagnostics <- bind_rows(
    county_join_diagnostics,
    tibble(
      dataset = "ACS",
      source_rows = nrow(acs_join_result),
      matched_counties = sum(matched, na.rm = TRUE),
      unmatched_rows = sum(!matched, na.rm = TRUE),
      match_rate = if (length(matched)) mean(matched, na.rm = TRUE) else NA_real_
    )
  )
}

if (exists("hrsa_nc") && !is.null(hrsa_nc) && nrow(hrsa_nc) > 0 &&
    exists("selected_boundaries") && map_geography == "county") {
  hrsa_join_result <- hrsa_nc |>
    mutate(county_join = clean_boundary_name(county_name))

  matched <- hrsa_join_result$county_join %in% selected_boundaries$county_join
  county_join_diagnostics <- bind_rows(
    county_join_diagnostics,
    tibble(
      dataset = "HRSA",
      source_rows = nrow(hrsa_join_result),
      matched_counties = sum(matched, na.rm = TRUE),
      unmatched_rows = sum(!matched, na.rm = TRUE),
      match_rate = if (length(matched)) mean(matched, na.rm = TRUE) else NA_real_
    )
  )
}

county_join_diagnostics |>
  mutate(match_rate = percent(match_rate, accuracy = 0.1)) |>
  DT::datatable(
    rownames = FALSE,
    options = list(dom = "t", pageLength = 10),
    class = "cell-border stripe hover"
  )
```



# Diagnostic summary
```{r diagnostic-summary, results='asis'}
cat("## Diagnostic render summary\n\n")

if (length(diagnostic_errors) == 0L) {
  cat("No R chunk errors were captured by knitr.\n")
} else {
  cat(sprintf("Captured %d R chunk error(s).\n\n", length(diagnostic_errors)))
  for (i in seq_along(diagnostic_errors)) {
    e <- diagnostic_errors[[i]]
    cat(sprintf("### Error %d: `%s`\n\n", i, e$label))
    cat("```text\n")
    cat(paste(e$message, collapse = "\n"))
    cat("\n```\n\n")
  }
}

if (length(diagnostic_warnings) > 0L) {
  cat(sprintf("### Warnings captured: %d\n\n", length(diagnostic_warnings)))
  for (i in seq_along(diagnostic_warnings)) {
    w <- diagnostic_warnings[[i]]
    cat(sprintf("- `%s`: %s\n", w$label, paste(w$message, collapse = " ")))
  }
}
```
