---
title: "extra"
output: html_document
---

---

# Part 2. NC DHHS Match Screening

Part 1 describes the whole file. Everything below narrows to opportunities that
**NC DHHS could realistically submit**, using the thresholds set in the config chunk.

> **Active screen.** Reference date `r format(ref_date, "%B %d, %Y")`.
> `r lbl_agency`. `r lbl_forecast`. `r lbl_runway`. `r lbl_eligible`.
> `r lbl_research`. `r lbl_restricted`.

```{r dhhs_subset, include = FALSE}
dhhs_grants <- if (is_unset(target_agencies)) {
  clean_grants_data
} else {
  clean_grants_data |> filter(top_level_agency_name %in% target_agencies)
}

n_dhhs <- nrow(dhhs_grants)
```

## Criterion Flags

Each rule below is stored as its own column. No rows are deleted at this stage, so
every exclusion can be audited later.

```{r flag_rows}
dhhs_flagged <- dhhs_grants |>
  mutate(
    elig_text = coalesce(applicant_eligibility_description, ""),
    search_text = paste(opportunity_title, opportunity_number),
    is_posted = if (keep_forecasts) TRUE else !is_forecast,
    has_runway = if (is.null(earliest_close)) {
      TRUE
    } else {
      !is.na(close_date) & close_date >= earliest_close
    },
    state_eligible = if (is_unset(required_applicant_types)) {
      TRUE
    } else {
      map_lgl(applicant_types, ~ any(required_applicant_types %in% .x))
    },
    flag_nih_agency = str_detect(coalesce(agency_code, ""), research_agency_regex),
    flag_research_number = str_detect(coalesce(opportunity_number, ""), research_number_regex),
    flag_activity_code = str_detect(search_text, research_activity_regex),
    flag_research_category = map_lgl(funding_categories, ~ research_category %in% .x),
    is_research_mechanism = flag_nih_agency | flag_research_number |
      flag_activity_code | flag_research_category,
    is_restricted = str_detect(elig_text, restricted_regex),
    drop_research = apply_research_rule & is_research_mechanism,
    drop_restricted = apply_restricted_rule & is_restricted,
    passes_screen = is_posted & has_runway & state_eligible &
      !drop_research & !drop_restricted
  ) |>
  select(-search_text)

flag_summary <- tibble(
  label = c(lbl_forecast, lbl_runway, lbl_eligible, lbl_research, lbl_restricted,
            "Passes all active screens"),
  n_true = c(
    sum(dhhs_flagged$is_posted, na.rm = TRUE),
    sum(dhhs_flagged$has_runway, na.rm = TRUE),
    sum(dhhs_flagged$state_eligible, na.rm = TRUE),
    sum(!dhhs_flagged$drop_research, na.rm = TRUE),
    sum(!dhhs_flagged$drop_restricted, na.rm = TRUE),
    sum(dhhs_flagged$passes_screen, na.rm = TRUE)
  )
) |>
  mutate(share = n_true / n_dhhs)

ggplot(flag_summary, aes(x = reorder(label, n_true), y = n_true)) +
  geom_col(fill = pal$primary, width = 0.7) +
  geom_text(aes(label = paste0(n_true, " (", percent(share, accuracy = 1), ")")),
            hjust = -0.1, size = 3.5, color = "#333333") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "How Many Opportunities Satisfy Each Criterion",
    subtitle = paste0("Out of ", n_dhhs, " opportunities after the agency filter"),
    x = NULL,
    y = "Opportunities surviving this rule on its own",
    caption = "A disabled rule keeps every row, so its bar equals the starting count."
  ) +
  theme(panel.grid.major.y = element_blank())
```

## Screening Funnel

```{r funnel, fig.width=10, fig.height=6}
step_agency <- dhhs_flagged
step_posted <- step_agency |> filter(is_posted)
step_runway <- step_posted |> filter(has_runway)
step_eligible <- step_runway |> filter(state_eligible)
step_nonresearch <- step_eligible |> filter(!drop_research)
step_open_comp <- step_nonresearch |> filter(!drop_restricted)

screened_grants <- step_open_comp

stage_labels <- c("All opportunities in file", lbl_agency, lbl_forecast,
                  lbl_runway, lbl_eligible, lbl_research, lbl_restricted)

funnel_data <- tibble(
  stage = factor(stage_labels, levels = rev(stage_labels)),
  n = c(nrow(clean_grants_data), nrow(step_agency), nrow(step_posted),
        nrow(step_runway), nrow(step_eligible), nrow(step_nonresearch),
        nrow(step_open_comp))
) |>
  mutate(dropped = lag(n) - n)

ggplot(funnel_data, aes(x = stage, y = n)) +
  geom_col(fill = pal$primary, width = 0.72) +
  geom_text(
    aes(label = case_when(
      is.na(dropped) ~ as.character(n),
      dropped == 0 ~ paste0(n, "  (no change)"),
      TRUE ~ paste0(n, "  (-", dropped, ")")
    )),
    hjust = -0.1, size = 3.6, color = "#333333"
  ) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "From Every Federal Posting to a Submittable Shortlist",
    subtitle = "Each bar applies one additional screening rule to the rows above it",
    x = NULL,
    y = "Opportunities remaining",
    caption = "Numbers in parentheses show how many opportunities that rule removed."
  ) +
  theme(panel.grid.major.y = element_blank())
```

`r nrow(screened_grants)` opportunities survive every active screen out of `r nrow(clean_grants_data)` in the source file.

```{r screened_table}
shortlist_flat <- step_eligible |>
  transmute(
    key = opportunity_id,
    title = opportunity_title,
    agency = agency_name,
    close_date,
    days_to_close = as.integer(close_date - ref_date),
    award_ceiling_num = coalesce(award_ceiling, 0),
    award_ceiling_fmt = if_else(!is.na(award_ceiling), dollar(award_ceiling), "N/A"),
    cost_share = if_else(coalesce(is_cost_sharing, FALSE), "Match required", "No match"),
    grant_type = case_when(
      drop_research ~ "Research mechanism",
      drop_restricted ~ "Restricted competition",
      TRUE ~ "Administrable by the department"
    ),
    link = paste0("<a href='", url, "' target='_blank'>Open</a>")
  ) |>
  filter(!is.na(days_to_close))

shared_grants <- SharedData$new(shortlist_flat, key = ~key, group = "grants")

type_colors <- c(
  "Administrable by the department" = pal$primary,
  "Research mechanism" = pal$accent,
  "Restricted competition" = pal$secondary
)

count_plot <- plot_ly(
  shared_grants,
  y = ~grant_type,
  color = ~grant_type,
  colors = type_colors,
  type = "histogram",
  orientation = "h",
  hovertemplate = "%{x} opportunities<extra></extra>"
) |>
  layout(
    title = list(text = "Opportunities Matching Your Filters", x = 0),
    xaxis = list(title = "Count"),
    yaxis = list(title = ""),
    showlegend = FALSE,
    margin = list(l = 10)
  ) |>
  config(displayModeBar = FALSE)

scatter_plot <- plot_ly(
  shared_grants,
  x = ~days_to_close,
  y = ~award_ceiling_num,
  color = ~grant_type,
  colors = type_colors,
  type = "scatter",
  mode = "markers",
  marker = list(size = 10, opacity = 0.75, line = list(width = 0.5, color = "white")),
  hoverinfo = "text",
  text = ~paste0(
    "<b>", title, "</b><br>", agency,
    "<br>Closes: ", close_date, " (", days_to_close, " days out)",
    "<br>Ceiling: ", award_ceiling_fmt,
    "<br>", cost_share
  )
) |>
  layout(
    title = list(text = "Deadline Runway vs Award Size", x = 0),
    xaxis = list(title = "Days from reference date to deadline"),
    yaxis = list(title = "Award ceiling ($, log scale)", type = "log"),
    legend = list(orientation = "h", y = -0.25)
  ) |>
  config(displayModeBar = FALSE)

bscols(
  widths = c(3, 9),
  list(
    filter_slider("runway", "Minimum days until deadline",
                  shared_grants, ~days_to_close, min = 0, step = 15),
    filter_slider("ceiling", "Minimum award ceiling ($)",
                  shared_grants, ~award_ceiling_num, step = 50000),
    filter_checkbox("type", "Opportunity type", shared_grants, ~grant_type),
    filter_checkbox("match", "Cost share", shared_grants, ~cost_share),
    filter_select("agency_pick", "Issuing office", shared_grants, ~agency)
  ),
  list(count_plot, scatter_plot)
)

datatable(
  shared_grants,
  escape = FALSE,
  rownames = FALSE,
  filter = "top",
  colnames = c("Key", "Title", "Agency", "Closes", "Days out", "Ceiling num",
               "Ceiling", "Cost share", "Type", "Link"),
  options = list(
    pageLength = 10,
    scrollX = TRUE,
    searchHighlight = TRUE,
    dom = "itfpt",
    columnDefs = list(list(visible = FALSE, targets = c(0, 5)))
  ),
  class = "cell-border stripe hover"
)
```

## Exclusion Audit

The two judgment-based rules are the research mechanism filter and the restricted
competition filter. Everything they removed is listed here so the tradeoff is visible
rather than hidden inside the ranking. If either rule is disabled in the config chunk,
it contributes nothing to this section.

```{r exclusion_audit, fig.width=10, fig.height=5}
excluded <- step_eligible |>
  filter(drop_research | drop_restricted)

exclusion_reasons <- tibble(
  signal = c("NIH issuing agency", "PA / PAR / PD / RFA number",
             "NIH activity code in title", "Research funding category",
             "Restricted or single source"),
  n = c(
    sum(excluded$flag_nih_agency & excluded$drop_research, na.rm = TRUE),
    sum(excluded$flag_research_number & excluded$drop_research, na.rm = TRUE),
    sum(excluded$flag_activity_code & excluded$drop_research, na.rm = TRUE),
    sum(excluded$flag_research_category & excluded$drop_research, na.rm = TRUE),
    sum(excluded$drop_restricted, na.rm = TRUE)
  )
)

ggplot(exclusion_reasons, aes(x = reorder(signal, n), y = n)) +
  geom_col(fill = pal$accent, width = 0.7) +
  geom_text(aes(label = n), hjust = -0.15, size = 3.6, color = "#333333") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Which Signal Removed Each Excluded Opportunity",
    subtitle = paste0(nrow(excluded), " eligible, still-open opportunities were excluded"),
    x = NULL,
    y = "Opportunities flagged by this signal",
    caption = "Signals overlap: one opportunity can be flagged by more than one rule."
  ) +
  theme(panel.grid.major.y = element_blank())
```

```{r exclusion_table}
excluded |>
  transmute(
    opportunity_title,
    agency_name,
    close_date,
    award_ceiling = if_else(!is.na(award_ceiling), dollar(award_ceiling), "N/A"),
    exclusion_reasons = paste(
      if_else(flag_nih_agency & drop_research, "NIH agency", ""),
      if_else(flag_research_number & drop_research, "research number", ""),
      if_else(flag_activity_code & drop_research, "activity code", ""),
      if_else(flag_research_category & drop_research, "research category", ""),
      if_else(drop_restricted, "restricted competition", ""),
      sep = "; "
    ) |>
      str_replace_all("(; )+", "; ") |>
      str_remove("^; ") |>
      str_remove("; $"),
    url
  ) |>
  arrange(desc(close_date)) |>
  make_dt(caption = "Every opportunity removed by the research and restricted competition rules.")
```