---
title: "Solve-A-Thon 2026: NC DHHS Federal Grant Opportunity Analysis"
author: "Kent Lee"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: cosmo
    smooth-scroll: true
    page-layout: full
execute:
  warning: false
  message: false
  echo: false
params:
  project_dir: "/Users/kentalee/Documents/KALProjects/UNCProjects/SolveAThon_2026"
  data_dir: "data"
  results_dir: "outs"
  cache_dir: "cache"
  grants_file: "grants-search-202608182008.csv"
  census_api_key: ""
  ref_date: "2026-09-30"
  min_runway_days: 45
  award_floor_usd: 250000
  target_agencies: "DHHS"
  nc_fips: "37"
---

<style>
:root {
  --blue:#2166AC; --blue-light:#D9EAF7; --purple:#756BB1;
  --orange:#E08214; --red:#B2182B; --green:#1B7837;
  --gray:#667085; --dark:#1D2939; --light:#F4F6F8;
}
body { color:var(--dark); }
h1,h2 { border-bottom:2px solid var(--blue-light); padding-bottom:.35rem; margin-top:2rem; }
h3 { margin-top:1.5rem; }
.kpi-grid { display:grid; grid-template-columns:repeat(auto-fit,minmax(180px,1fr)); gap:14px; margin:18px 0 28px 0; }
.kpi-card { background:white; border:1px solid #E4E7EC; border-left:5px solid var(--blue); border-radius:8px; padding:16px; min-height:100px; box-shadow:0 2px 7px rgba(16,24,40,.06); }
.kpi-card.warning { border-left-color:var(--orange); }
.kpi-card.danger  { border-left-color:var(--red); }
.kpi-card.success { border-left-color:var(--green); }
.kpi-label { color:var(--gray); font-size:.82rem; text-transform:uppercase; letter-spacing:.04em; }
.kpi-value { color:var(--dark); font-size:1.8rem; font-weight:700; margin-top:5px; }
.kpi-note  { color:var(--gray); font-size:.78rem; margin-top:5px; }
.callout { background:#F8FAFC; border-left:5px solid var(--blue); padding:14px 18px; margin:18px 0; border-radius:4px; }
.callout-warning { border-left-color:var(--orange); background:#FFFAEB; }
.callout-green   { border-left-color:var(--green);  background:#EDF7ED; }
.data-source-box { background:#F0F7F0; border-left:5px solid var(--green); padding:12px 16px; margin:14px 0; border-radius:4px; font-size:.88em; }
.dataTables_wrapper { font-size:.88em; margin-top:15px; margin-bottom:30px; }
table.dataTable thead th { background-color:#EEF4FA; }
pre { border-radius:6px; }
</style>

```{r setup}
knitr::opts_chunk$set(
  echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE,
  fig.align="center", dpi=150, out.width="100%"
)
project_dir <- params$project_dir
data_dir    <- file.path(project_dir, params$data_dir)
results_dir <- file.path(project_dir, params$results_dir)
cache_dir   <- file.path(data_dir,    params$cache_dir)
knitr::opts_knit$set(root.dir = project_dir)
for (d in c(results_dir, cache_dir,
            file.path(cache_dir, "api"),
            file.path(cache_dir, "llm"),
            file.path(data_dir, "raw", "boundaries"),
            file.path(data_dir, "raw", "acs"),
            file.path(data_dir, "raw", "hrsa"),
            file.path(data_dir, "raw", "cdc_places"),
            file.path(data_dir, "raw", "cdc_wonder"),
            file.path(data_dir, "raw", "fema"),
            file.path(data_dir, "raw", "usaspending"),
            file.path(data_dir, "raw", "samhsa"),
            file.path(data_dir, "processed"))) {
  dir.create(d, showWarnings=FALSE, recursive=TRUE)
}
options(bitmapType="cairo", tigris_use_cache=TRUE)
```

```{r libraries}
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(stringr)
  library(lubridate); library(janitor); library(purrr)
  library(ggplot2); library(scales); library(DT)
  library(plotly); library(forcats); library(htmltools)
  # Spatial and census
  library(sf); library(tigris); library(tidycensus)
  # API calls
  library(httr2); library(jsonlite)
})
```

```{r config}
ref_date_input                  <- params$ref_date
min_runway_days                 <- params$min_runway_days
award_floor_usd                 <- params$award_floor_usd
target_agencies                 <- params$target_agencies
include_forecasts               <- FALSE
exclude_research_mechanisms     <- TRUE
exclude_restricted_competitions <- TRUE
keep_forecasts                  <- include_forecasts

ref_date       <- ymd(ref_date_input)
earliest_close <- ref_date + days(min_runway_days)

nc_fips        <- params$nc_fips
peer_fips      <- c("13","51","45","47")      # GA, VA, SC, TN
peer_abbr      <- c("GA","VA","SC","TN","NC")

# Census API key — set in params or call census_api_key() interactively
if (nchar(params$census_api_key) > 0) {
  tidycensus::census_api_key(params$census_api_key, install=FALSE)
}

pal <- list(
  blue="#2166AC", blue_light="#D9EAF7", purple="#756BB1",
  orange="#E08214", red="#B2182B", green="#1B7837",
  gray="#667085", dark="#1D2939", light="#F4F6F8"
)

theme_g <- function(base_size=12) {
  theme_minimal(base_size=base_size) +
    theme(
      plot.title    = element_text(face="bold", size=base_size+3, color=pal$dark),
      plot.subtitle = element_text(color=pal$gray, margin=margin(b=12)),
      plot.caption  = element_text(color=pal$gray, size=base_size-2, hjust=0),
      axis.title    = element_text(color=pal$dark),
      axis.text     = element_text(color="#475467"),
      panel.grid.minor   = element_blank(),
      panel.grid.major.y = element_blank(),
      plot.margin   = margin(12,20,12,12)
    )
}
theme_set(theme_g())
```

```{r helper-functions, include=FALSE}
is_unset    <- function(x) is.null(x) || length(x)==0L || all(is.na(x))
safe_false  <- function(x) dplyr::coalesce(as.logical(x), FALSE)

# Cache helper — returns cached object or calls fetch_fn() and saves result
get_cached <- function(path, fetch_fn) {
  if (file.exists(path)) return(readRDS(path))
  result <- tryCatch(fetch_fn(), error = function(e) {
    message("Fetch failed: ", conditionMessage(e)); NULL
  })
  if (!is.null(result)) saveRDS(result, path)
  result
}
```

```{r load-and-clean}
input_file <- file.path(data_dir, "raw", "grants", params$grants_file)
# Also accept file directly in data_dir for backward compatibility
if (!file.exists(input_file)) input_file <- file.path(data_dir, params$grants_file)
if (!file.exists(input_file)) stop("Grants file not found: ", input_file)

grants_raw <- readr::read_csv(input_file, show_col_types=FALSE,
                               na=c("","NA","N/A","NULL")) |>
  janitor::clean_names()

# ── Date parser: base-R only (lubridate parse_date_time exceeds PCRE limits) ──
parse_date_safe <- function(x) {
  x <- trimws(as.character(x))
  x[x %in% c("","NA","N/A","NULL","null")] <- NA_character_
  x <- sub("T.*$", "", x, perl=TRUE)   # strip ISO 8601 time component
  x <- sub(" .*$", "", x, perl=TRUE)   # strip space-separated time
  result <- suppressWarnings(as.Date(x))
  for (fmt in c("%m/%d/%Y", "%m/%d/%y", "%Y/%m/%d")) {
    miss <- is.na(result) & !is.na(x)
    if (!any(miss)) break
    result[miss] <- suppressWarnings(as.Date(x[miss], format=fmt))
  }
  result
}

parse_numeric_safe <- function(x) {
  suppressWarnings(as.numeric(str_replace_all(as.character(x), "[$,]", "")))
}

# Compile patterns once
research_pattern <- paste(c(
  "investigator initiated","investigator-initiated","research project grant",
  "clinical trial","research grant","\\br01\\b","\\br03\\b","\\br21\\b",
  "\\br34\\b","\\bu01\\b","\\bp01\\b","\\bp20\\b","\\bp30\\b","\\bp50\\b",
  "\\bk01\\b","\\bk08\\b","\\bk23\\b","\\bt32\\b","\\bf31\\b","\\bf32\\b"),
  collapse="|")

restricted_pattern <- regex(paste(c(
  "only the following applicant","only the following entit",
  "single source","sole source","invited applicant","not a competitive"),
  collapse="|"), ignore_case=TRUE)

date_cols <- intersect(
  c("post_date","close_date","archive_date","forecasted_post_date",
    "forecasted_close_date","forecasted_award_date","forecasted_project_start_date"),
  names(grants_raw))
num_cols  <- intersect(
  c("award_floor","award_ceiling","estimated_total_program_funding",
    "expected_number_of_awards"), names(grants_raw))

grants <- grants_raw |>
  mutate(across(all_of(date_cols), parse_date_safe),
         across(all_of(num_cols),  parse_numeric_safe)) |>
  mutate(
    close_date  = if_else(close_date == as.Date("2099-01-01"), as.Date(NA), close_date),
    is_forecast = safe_false(is_forecast),
    across(any_of(c("opportunity_title","agency_name","top_level_agency_name",
                    "summary_description","applicant_types","funding_instruments",
                    "funding_categories","applicant_eligibility_description")),
           ~ str_squish(as.character(.x))),

    status_group = case_when(
      is_forecast                        ~ "Forecasted",
      opportunity_status == "forecasted" ~ "Forecasted",
      TRUE                               ~ "Posted"),

    agency_group = case_when(
      top_level_agency_name == "Department of Health and Human Services" ~ "DHHS",
      top_level_agency_name == "U.S. National Science Foundation"        ~ "NSF",
      top_level_agency_name == "Department of Defense"                   ~ "DOD",
      top_level_agency_name == "Department of Agriculture"               ~ "USDA",
      top_level_agency_name == "Department of Labor"                     ~ "DOL",
      top_level_agency_name == "Department of Homeland Security"         ~ "DHS",
      top_level_agency_name == "Department of Transportation"            ~ "DOT",
      TRUE                                                                ~ top_level_agency_name),

    applicant_types_text = str_to_lower(coalesce(applicant_types,"")),
    funding_text = str_to_lower(str_c(
      coalesce(opportunity_title,""), coalesce(summary_description,""),
      coalesce(applicant_eligibility_description,""),
      coalesce(category,""), coalesce(category_explanation,""), sep=" | ")),

    state_eligible            = str_detect(applicant_types_text, fixed("state_governments")),
    is_nih = str_detect(str_to_lower(str_c(
      coalesce(agency_name,""), coalesce(top_level_agency_name,""),
      coalesce(opportunity_title,""), sep=" | ")),
      "\\bnih\\b|national institutes of health|hhs-nih"),
    is_research_mechanism     = str_detect(funding_text, research_pattern),
    is_restricted_competition = str_detect(funding_text, restricted_pattern),

    days_to_close = as.numeric(close_date - ref_date),
    has_sufficient_runway = is.na(close_date) | days_to_close >= min_runway_days,
    closes_soon = !is.na(days_to_close) & days_to_close >= 0 & days_to_close <= 30,

    close_window = case_when(
      is.na(close_date)    ~ "No stated deadline",
      days_to_close < 0    ~ "Already closed",
      days_to_close <= 30  ~ "0-30 days",
      days_to_close <= 60  ~ "31-60 days",
      days_to_close <= 120 ~ "61-120 days",
      TRUE                 ~ "More than 120 days"),

    award_size = coalesce(award_ceiling, award_floor),
    award_size_group = case_when(
      is.na(award_size)     ~ "Not reported",
      award_size < 250000   ~ "Under $250K",
      award_size < 1000000  ~ "$250K-$1M",
      award_size < 5000000  ~ "$1M-$5M",
      award_size < 25000000 ~ "$5M-$25M",
      TRUE                  ~ "$25M+")
  )
```

## Data quality

```{r data-quality}
tibble(
  measure = c("Rows","Columns","Posted","Forecasted",
              "With close date","With award ceiling",
              "With applicant eligibility","With summary"),
  value   = c(nrow(grants), ncol(grants),
              sum(grants$status_group=="Posted",    na.rm=TRUE),
              sum(grants$status_group=="Forecasted",na.rm=TRUE),
              sum(!is.na(grants$close_date)),
              sum(!is.na(grants$award_ceiling)),
              sum(!is.na(grants$applicant_eligibility_description)),
              sum(!is.na(grants$summary_description)))
) |>
  DT::datatable(rownames=FALSE, options=list(dom="t", pageLength=10))
```
# 1. Executive Summary

```{r exec-summary, results='asis'}
candidate_count <- grants |>
  filter(status_group=="Posted", state_eligible, !is_research_mechanism,
         !is_restricted_competition,
         is.na(close_date) | close_date >= earliest_close,
         is.na(award_size)  | award_size  >= award_floor_usd) |> nrow()

cat('<div class="kpi-grid">',
  '<div class="kpi-card"><div class="kpi-label">Total opportunities</div>',
  '<div class="kpi-value">',comma(nrow(grants)),'</div>',
  '<div class="kpi-note">Posted and forecasted</div></div>',
  '<div class="kpi-card"><div class="kpi-label">Posted</div>',
  '<div class="kpi-value">',comma(sum(grants$status_group=="Posted",na.rm=TRUE)),'</div>',
  '<div class="kpi-note">In snapshot</div></div>',
  '<div class="kpi-card success"><div class="kpi-label">State-eligible</div>',
  '<div class="kpi-value">',comma(sum(grants$state_eligible,na.rm=TRUE)),'</div>',
  '<div class="kpi-note">Initial screen only</div></div>',
  '<div class="kpi-card danger"><div class="kpi-label">NIH tagged</div>',
  '<div class="kpi-value">',comma(sum(grants$is_nih,na.rm=TRUE)),'</div>',
  '<div class="kpi-note">Research-noise source</div></div>',
  '<div class="kpi-card warning"><div class="kpi-label">Screened candidates</div>',
  '<div class="kpi-value">',comma(candidate_count),'</div>',
  '<div class="kpi-note">After all filters</div></div>',
  '</div>')
```

<div class="callout">
This dashboard is an initial screening view. Eligibility tags are broad; the
research-mechanism screen is text-based and must be validated on a random sample
before excluding any opportunity from a final shortlist.
</div>

# 2. Screening Funnel

Each stage is applied only to rows that passed every previous stage.

```{r funnel-prep}
agency_pass     <- if (is_unset(target_agencies)) rep(TRUE,nrow(grants)) else
                   grants$agency_group %in% target_agencies | grants$top_level_agency_name %in% target_agencies
status_pass     <- if (keep_forecasts) rep(TRUE,nrow(grants)) else grants$status_group=="Posted"
eligibility_pass<- safe_false(grants$state_eligible)
research_pass   <- if (isTRUE(exclude_research_mechanisms))  !safe_false(grants$is_research_mechanism)  else rep(TRUE,nrow(grants))
restricted_pass <- if (isTRUE(exclude_restricted_competitions)) !safe_false(grants$is_restricted_competition) else rep(TRUE,nrow(grants))
runway_pass     <- if (is.null(min_runway_days)) rep(TRUE,nrow(grants)) else
                   !is.na(grants$close_date) & !is.na(grants$days_to_close) & grants$days_to_close >= min_runway_days
award_pass      <- if (is.null(award_floor_usd)) rep(TRUE,nrow(grants)) else
                   is.na(grants$award_size) | grants$award_size >= award_floor_usd

funnel_rules <- list(
  `All opportunities`      = rep(TRUE,nrow(grants)),
  `Selected agency`        = agency_pass,
  `Posted`                 = status_pass,
  `State eligible`         = eligibility_pass,
  `Not research mechanism` = research_pass,
  `Open competition`       = restricted_pass,
  `Enough runway`          = runway_pass,
  `Above award threshold`  = award_pass)

survivors <- rep(TRUE, nrow(grants))
funnel_data <- purrr::imap_dfr(funnel_rules, function(rule, name) {
  survivors <<- survivors & rule
  tibble(stage=name, n=sum(survivors, na.rm=TRUE))
}) |>
  mutate(
    dropped = c(0, head(n,-1) - tail(n,-1)),
    pct_all = n / first(n),
    stage   = factor(stage, levels=rev(stage)),
    label   = paste0(comma(n)," (",percent(pct_all, accuracy=0.1),")")
  )
funnel_data
```

```{r funnel-chart, fig.width=11, fig.height=7}
ggplot(funnel_data, aes(x=n, y=stage)) +
  geom_col(fill=pal$blue, width=.68) +
  geom_text(aes(label=label), hjust=-.08, size=4, color=pal$dark) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.18))) +
  labs(title="Sequential Screening Funnel",
       subtitle=paste("Reference date:", format(ref_date,"%B %d, %Y"),
                      "| Minimum runway:", min_runway_days, "days"),
       x="Opportunities remaining", y=NULL) +
  theme(axis.text.y=element_text(face="bold"), panel.grid.major.y=element_blank())
```

```{r funnel-table}
funnel_data |>
  mutate(across(c(n,dropped), comma),
         pct_all=percent(pct_all, accuracy=0.1),
         stage=as.character(stage)) |>
  select(stage, n, dropped, pct_all) |>
  DT::datatable(rownames=FALSE,
    colnames=c("Stage","Remaining","Removed at stage","% of original"),
    options=list(dom="t", pageLength=10, ordering=FALSE),
    class="cell-border stripe hover")
```

# 3. Opportunity Volume by Agency

```{r agency-vol, fig.width=11, fig.height=8}
agency_dat <- grants |>
  count(agency_group, status_group, name="n") |>
  group_by(agency_group) |> mutate(total=sum(n)) |> ungroup() |>
  filter(total >= 5) |>
  mutate(agency_group=fct_reorder(agency_group, total))

ggplot(agency_dat, aes(x=n, y=agency_group, fill=status_group)) +
  geom_col(position="stack", width=.72) +
  geom_text(data=agency_dat |> group_by(agency_group) |>
              summarise(total=first(total),.groups="drop"),
            aes(x=total, y=agency_group, label=comma(total)),
            inherit.aes=FALSE, hjust=-.12, size=3.5, color=pal$dark) +
  scale_fill_manual(values=c(Posted=pal$blue, Forecasted=pal$purple)) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.14))) +
  labs(title="Federal Opportunity Volume by Agency",
       subtitle="Agencies with < 5 opportunities omitted",
       x="Opportunities", y=NULL, fill=NULL) +
  theme(legend.position="top", panel.grid.major.y=element_blank())
```

# 4. Application Deadline Urgency

```{r deadline-urgency-data}
deadline_dat <- grants |>
  filter(status_group=="Posted", !is.na(close_date), close_date >= ref_date) |>
  mutate(close_window = factor(case_when(
    days_to_close <= 30  ~ "0-30 days",
    days_to_close <= 60  ~ "31-60 days",
    days_to_close <= 120 ~ "61-120 days",
    TRUE                 ~ "More than 120 days"),
    levels=c("0-30 days","31-60 days","61-120 days","More than 120 days"))) |>
  count(close_window)
```

```{r deadline-urgency-plot, fig.width=10, fig.height=6}
ggplot(deadline_dat, aes(x=close_window, y=n, fill=close_window)) +
  geom_col(width=.7, show.legend=FALSE) +
  geom_text(aes(label=comma(n)), vjust=-.35, size=4) +
  scale_fill_manual(values=c("0-30 days"=pal$red,"31-60 days"=pal$orange,
                              "61-120 days"=pal$blue,"More than 120 days"=pal$green)) +
  scale_y_continuous(labels=comma, expand=expansion(mult=c(0,.12))) +
  labs(title="Posted Opportunities by Remaining Application Window",
       subtitle=paste("From", format(ref_date,"%B %d, %Y")),
       x=NULL, y="Opportunities") +
  theme(panel.grid.major.x=element_blank())
```

# 5. Award Amount Distribution

Award values span orders of magnitude; the x-axis is logarithmic. A national program
ceiling is **not** the amount North Carolina would receive — it is a total across all
awardees nationwide.

```{r award-dist, fig.width=11, fig.height=6}
award_dist_dat <- grants |> filter(status_group=="Posted", !is.na(award_size), award_size > 0)

ggplot(award_dist_dat, aes(x=award_size)) +
  geom_histogram(bins=35, fill=pal$blue, color="white", alpha=.9) +
  geom_vline(xintercept=award_floor_usd, color=pal$red, linetype="dashed", linewidth=1) +
  annotate("text", x=award_floor_usd, y=Inf,
           label=paste0("Threshold: ",dollar(award_floor_usd)),
           hjust=-.05, vjust=1.5, color=pal$red, size=3.5) +
  scale_x_log10(labels=dollar_format(scale_cut=cut_short_scale())) +
  scale_y_continuous(labels=comma) +
  labs(title="Distribution of Reported Award Amounts (Posted)",
       subtitle="Uses award ceiling when available; award floor otherwise",
       x="Award amount (log scale)", y="Opportunities") +
  theme(panel.grid.minor.x=element_blank())
```

# 6. Funding Instruments

A **cooperative agreement** carries ongoing federal involvement and is an early
signal of higher administrative burden compared to a standard grant.

```{r instruments, fig.width=10, fig.height=6}
grants |>
  filter(status_group=="Posted") |>
  mutate(instrument = case_when(
    str_detect(str_to_lower(coalesce(funding_instruments,"")), "cooperative") ~ "Cooperative agreement",
    str_detect(str_to_lower(coalesce(funding_instruments,"")), "\\bgrant\\b")  ~ "Grant",
    str_detect(str_to_lower(coalesce(funding_instruments,"")), "procurement|contract") ~ "Procurement contract",
    str_detect(str_to_lower(coalesce(funding_instruments,"")), "other") ~ "Other",
    TRUE ~ "Not reported")) |>
  count(instrument, sort=TRUE) |>
  mutate(instrument=fct_reorder(instrument,n)) |>
  ggplot(aes(x=n, y=instrument, fill=instrument)) +
  geom_col(width=.7, show.legend=FALSE) +
  geom_text(aes(label=comma(n)), hjust=-.12, size=4) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.14))) +
  labs(title="Posted Opportunities by Funding Instrument",
       subtitle="Cooperative agreements imply ongoing federal oversight",
       x="Posted opportunities", y=NULL) +
  theme(panel.grid.major.y=element_blank())
```

# 7. State Eligibility and Research Noise

<div class="callout callout-green">
**What does "state_governments" eligibility mean?**
Grants.gov uses a controlled vocabulary for applicant types. The code
`state_governments` indicates that the posting explicitly lists state government
entities as eligible applicants. This is one of approximately 15 applicant-type
codes in the schema. **It does not mean the state is the sole or even primary
intended recipient.** Of 887 state-eligible rows in this snapshot, only 12 are
restricted to state governments alone. Most also include universities, hospitals,
nonprofits, tribes, and for-profits — making eligibility alone a very weak
discriminator. The research-mechanism screen below is needed to separate
administrative grants from investigator-initiated research.
</div>

```{r eligibility-research-data}
elig_res <- grants |>
  mutate(
    elig_grp     = if_else(safe_false(state_eligible), "State eligible","Not state eligible"),
    research_grp = if_else(safe_false(is_research_mechanism),"Research screen","Not research screen")
  ) |>
  count(elig_grp, research_grp) |>
  group_by(elig_grp) |> mutate(pct=n/sum(n)) |> ungroup()
```

```{r eligibility-research-plot, fig.width=10, fig.height=6}
ggplot(elig_res, aes(x=elig_grp, y=n, fill=research_grp)) +
  geom_col(position="stack", width=.68) +
  geom_text(aes(label=paste0(comma(n),"\n",percent(pct,accuracy=0.1))),
            position=position_stack(vjust=.5), color="white", fontface="bold", size=3.5) +
  scale_fill_manual(values=c("Research screen"=pal$red,"Not research screen"=pal$blue)) +
  scale_y_continuous(labels=comma) +
  labs(title="State Eligibility Does Not Identify Administrative Fit",
       subtitle="Research mechanism screen is preliminary and text-based only",
       x=NULL, y="Opportunities", fill=NULL) +
  theme(legend.position="top", panel.grid.major.x=element_blank())
```

<div class="callout callout-warning">
Validate the research screen on a random hand-labelled sample of at least 60
rows before using it to exclude opportunities from the final shortlist.
</div>

# 8. Shortlist After Screening

```{r shortlist-data}
shortlist <- grants |>
  mutate(passed = survivors) |>
  filter(passed) |>
  mutate(
    urgency = case_when(
      is.na(days_to_close) ~ "No stated deadline",
      days_to_close <= 30  ~ "Urgent: 0-30 days",
      days_to_close <= 60  ~ "Near-term: 31-60 days",
      days_to_close <= 120 ~ "Medium-term: 61-120 days",
      TRUE                 ~ "Longer runway"),
    award_display = case_when(
      !is.na(award_ceiling) ~ paste0("Ceiling: ",dollar(award_ceiling)),
      !is.na(award_floor)   ~ paste0("Floor: ",  dollar(award_floor)),
      TRUE                  ~ "Not reported"),
    desc_trunc = str_trunc(coalesce(summary_description,"No summary"), 280)
  ) |>
  arrange(days_to_close, desc(award_size)) |>
  select(opportunity_number, opportunity_title, agency_group,
         urgency, close_date, days_to_close, award_display,
         state_eligible, is_nih, is_research_mechanism,
         opportunity_assistance_listings, desc_trunc, additional_info_url)
```

```{r shortlist-table}
shortlist |>
  mutate(close_date=if_else(is.na(close_date),"No deadline",format(close_date,"%Y-%m-%d"))) |>
  DT::datatable(rownames=FALSE, escape=FALSE, filter="top",
    extensions=c("Buttons","FixedHeader"),
    options=list(pageLength=15, scrollX=TRUE, fixedHeader=TRUE,
                 dom="Bfrtip", buttons=c("copy","csv","excel"),
                 searchHighlight=TRUE),
    class="cell-border stripe hover") |>
  DT::formatStyle("urgency", backgroundColor=DT::styleEqual(
    c("Urgent: 0-30 days","Near-term: 31-60 days",
      "Medium-term: 61-120 days","Longer runway","No stated deadline"),
    c("#FDE2E2","#FFF0D5","#E7F0FA","#E6F4EA","#F2F4F7")))
```

# 9. Opportunities Closing Soon

All posted opportunities closing within 30 days — regardless of whether they
pass the full screening funnel.

```{r urgent-table}
grants |>
  filter(status_group=="Posted", !is.na(close_date),
         close_date >= ref_date, days_to_close <= 30) |>
  mutate(
    award_display = case_when(
      !is.na(award_ceiling) ~ dollar(award_ceiling),
      !is.na(award_floor)   ~ dollar(award_floor),
      TRUE ~ "Not reported"),
    desc_trunc = str_trunc(coalesce(summary_description,"No summary"), 220)
  ) |>
  arrange(days_to_close) |>
  select(opportunity_number, opportunity_title, agency_group,
         close_date, days_to_close, state_eligible, is_nih,
         is_research_mechanism, award_display, desc_trunc) |>
  DT::datatable(rownames=FALSE, filter="top",
    extensions=c("Buttons","FixedHeader"),
    options=list(pageLength=15, scrollX=TRUE, fixedHeader=TRUE,
                 dom="Bfrtip", buttons=c("copy","csv","excel")),
    class="cell-border stripe hover")
```

# 10. Close-Date vs Award Size

Outliers above the 95th percentile of award size (mostly large national NIH and
SAMHSA program totals) are removed to keep the scale informative for
state-administrable grants.

```{r runway-award-data}
runway_award_dat <- grants |>
  filter(status_group=="Posted", !is.na(close_date), close_date >= ref_date,
         !is.na(award_size), award_size > 0) |>
  mutate(
    elig_disp = if_else(safe_false(state_eligible),"State eligible","Not state eligible"),
    res_disp  = if_else(safe_false(is_research_mechanism),"Research screen","Not research screen")
  )

p95 <- quantile(runway_award_dat$award_size, .95, na.rm=TRUE)
runway_award_trim <- runway_award_dat |> filter(award_size <= p95)
```

```{r runway-award-plot, fig.width=11, fig.height=7}
ra_plot <- ggplot(runway_award_trim,
  aes(x=days_to_close, y=award_size,
      color=elig_disp, shape=res_disp, text=opportunity_title)) +
  geom_point(alpha=.7, size=2.5) +
  geom_vline(xintercept=min_runway_days,
             linetype="dashed", color=pal$red, linewidth=.9) +
  annotate("text", x=min_runway_days, y=Inf,
           label=paste0(min_runway_days,"-day runway"),
           hjust=1.05, vjust=1.5, color=pal$red, size=3.2) +
  scale_x_continuous(labels=comma) +
  scale_y_continuous(labels=dollar_format(scale_cut=cut_short_scale())) +
  scale_color_manual(values=c("State eligible"=pal$green,"Not state eligible"=pal$gray)) +
  scale_shape_manual(values=c("Research screen"=17,"Not research screen"=16)) +
  labs(title="Remaining Runway vs Award Size (outliers removed)",
       subtitle=paste0("Awards above the 95th percentile (>",dollar(round(p95))," ) excluded"),
       x="Days until close date", y="Award size", color=NULL, shape=NULL) +
  theme(legend.position="top")

ggplotly(ra_plot, tooltip=c("x","y","colour","shape","text"))
```

# 11. Most Frequent Assistance Listings

Assistance listings (CFDA numbers) are the join key for USAspending award history.
Programs appearing repeatedly in the snapshot are the highest-priority lookup targets.

```{r cfda-plot, fig.width=11, fig.height=9}
grants |>
  filter(!is.na(opportunity_assistance_listings),
         str_squish(opportunity_assistance_listings) != "") |>
  separate_rows(opportunity_assistance_listings, sep="\\s*;\\s*") |>
  mutate(opportunity_assistance_listings=str_squish(opportunity_assistance_listings)) |>
  filter(opportunity_assistance_listings != "") |>
  count(opportunity_assistance_listings, sort=TRUE, name="n") |>
  slice_head(n=25) |>
  mutate(opportunity_assistance_listings=fct_reorder(opportunity_assistance_listings,n)) |>
  ggplot(aes(x=n, y=opportunity_assistance_listings)) +
  geom_col(fill=pal$purple, width=.7) +
  geom_text(aes(label=comma(n)), hjust=-.12, size=3.3) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.15))) +
  labs(title="Most Frequent Assistance Listings (top 25)",
       subtitle="These CFDA numbers are the priority set for USAspending joins",
       x="Opportunities in snapshot", y=NULL) +
  theme(panel.grid.major.y=element_blank())
```

# 12. Preliminary Triage Quadrants

This view plots every posted opportunity with a stated deadline and reported award
amount against two observable axes: **days of runway** and **award size**. It is a
workload and prioritisation diagnostic — not a measure of need, fit, or probability
of winning.

**How to read the quadrants:**

| Quadrant | Description | Suggested action |
|---|---|---|
| Large award, sufficient runway | High potential value, time to prepare | Assign owner; begin fit assessment |
| Large award, short runway | High value but fast-closing | Leadership decision this week: pursue or formally decline |
| Smaller award, sufficient runway | Lower ceiling; adequate time | Route to relevant division for optional review |
| Smaller award, short runway | Limited time and ceiling | Only pursue if strong existing relationship with program |

Opportunities flagged green (initial fit screen: state-eligible and not a research
mechanism) are the primary triage targets. The 95th-percentile cap removes large
national-program totals that distort the scale.

```{r quadrant-data}
quad_dat <- grants |>
  filter(status_group=="Posted", !is.na(close_date), close_date >= ref_date,
         !is.na(award_size), award_size > 0) |>
  mutate(
    fit = case_when(
      safe_false(state_eligible) & !safe_false(is_research_mechanism) ~ "Initial fit screen",
      safe_false(state_eligible) &  safe_false(is_research_mechanism) ~ "Eligible; research screen",
     !safe_false(state_eligible)                                       ~ "Not state eligible"),
    quadrant = case_when(
      days_to_close <= min_runway_days & award_size >= award_floor_usd ~ "Large award, short runway",
      days_to_close >  min_runway_days & award_size >= award_floor_usd ~ "Large award, sufficient runway",
      days_to_close <= min_runway_days & award_size <  award_floor_usd ~ "Smaller award, short runway",
      TRUE                                                              ~ "Smaller award, sufficient runway")
  )

p95q <- quantile(quad_dat$award_size, .95, na.rm=TRUE)
quad_trim <- quad_dat |> filter(award_size <= p95q)
```

```{r quadrant-plot, fig.width=11, fig.height=7}
qp <- ggplot(quad_trim,
  aes(x=days_to_close, y=award_size,
      color=fit, shape=quadrant, text=opportunity_title)) +
  geom_point(alpha=.7, size=2.5) +
  geom_vline(xintercept=min_runway_days, linetype="dashed", linewidth=.9, color=pal$red) +
  geom_hline(yintercept=award_floor_usd,  linetype="dashed", linewidth=.9, color=pal$orange) +
  annotate("text", x=min_runway_days, y=Inf,
           label=paste0(min_runway_days,"-day runway"),
           hjust=1.05, vjust=1.5, color=pal$red, size=3.0) +
  annotate("text", x=Inf, y=award_floor_usd,
           label=paste0("Floor: ",dollar(award_floor_usd)),
           hjust=1.05, vjust=-.5, color=pal$orange, size=3.0) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(.02,.06))) +
  scale_y_continuous(labels=dollar_format(scale_cut=cut_short_scale())) +
  scale_color_manual(values=c("Initial fit screen"=pal$green,
                               "Eligible; research screen"=pal$red,
                               "Not state eligible"=pal$gray)) +
  labs(title="Preliminary Triage: Runway vs Award Size (outliers removed)",
       subtitle=paste("Reference date:", format(ref_date,"%B %d, %Y"),
                      "| 95th pct cap:", dollar(round(p95q))),
       x="Days until close date", y="Award size",
       color="Initial screen", shape="Quadrant") +
  theme(legend.position="bottom", legend.box="vertical")

ggplotly(qp, tooltip=c("x","y","colour","shape","text"))
```

```{r quadrant-counts}
quad_dat |> count(quadrant, fit, sort=TRUE) |>
  mutate(pct=percent(n/sum(n), accuracy=0.1)) |>
  DT::datatable(rownames=FALSE,
    colnames=c("Quadrant","Screen result","Opportunities","Share"),
    options=list(dom="t", pageLength=15),
    class="cell-border stripe hover")
```

<div class="callout callout-warning">
This chart does not measure need, capacity, or probability of success. It is a
workload-planning tool only. Do not use award ceilings as estimates of what North
Carolina would receive.
</div>

# 13. Forecasted Opportunity Pipeline

Forecasted opportunities are not open for application. They support calendar
planning and allow early internal review of likely requirements.

```{r forecast-pipeline, fig.width=11, fig.height=6}
grants |>
  filter(status_group=="Forecasted") |>
  mutate(
    fd = coalesce(forecasted_post_date, forecasted_close_date),
    fq = if_else(!is.na(fd), paste0(year(fd)," Q",quarter(fd)), "Date missing")
  ) |>
  count(fq) |>
  mutate(fq=factor(fq, levels=unique(fq))) |>
  ggplot(aes(x=fq, y=n)) +
  geom_col(fill=pal$purple, width=.7) +
  geom_text(aes(label=comma(n)), vjust=-.35, size=4) +
  scale_y_continuous(labels=comma, expand=expansion(mult=c(0,.12))) +
  labs(title="Forecasted Opportunity Pipeline by Quarter",
       subtitle="Forecasts support calendar planning before an opportunity posts",
       x="Expected posting quarter", y="Forecasted opportunities") +
  theme(axis.text.x=element_text(angle=35, hjust=1),
        panel.grid.major.x=element_blank())
```

# 14. NCDHHS-Relevant Domain Classification

Each opportunity may match more than one domain. Unmatched rows should be reviewed,
not assumed irrelevant.

```{r domain-patterns}
domain_patterns <- c(
  "Behavioral health and crisis"          = "behavioral health|mental health|suicide|988|crisis|psychiatric|trauma",
  "Opioid and substance use"              = "opioid|substance use|overdose|naloxone|addiction|recovery|syringe",
  "Maternal, infant, reproductive health" = "maternal|infant|pregnan|prenatal|reproductive|family planning|title x",
  "Child and family well-being"           = "child welfare|children|family|foster|early childhood|parenting|child care",
  "Food and nutrition security"           = "food security|nutrition|wic|snap|fns|hunger|food insecurity",
  "Health and direct-care workforce"      = "health workforce|workforce|nursing|direct care|recruitment|retention",
  "Rural access and telehealth"           = "rural|telehealth|telemedicine|underserved|health access",
  "Disaster recovery and resilience"      = "disaster|resilience|recovery|emergency|hurricane|flood|hazard"
)
```

```{r domain-classification}
domain_text_vec <- str_to_lower(str_c(
  coalesce(grants$opportunity_title,""),
  coalesce(grants$summary_description,""),
  coalesce(grants$applicant_eligibility_description,""),
  coalesce(grants$category,""), coalesce(grants$category_explanation,""),
  sep=" | "))

# FIX: use apply() — pmap_chr with select(.) fails inside mutate()
domain_flags <- as.data.frame(purrr::map_dfc(
  domain_patterns,
  ~ str_detect(domain_text_vec, regex(.x, ignore_case=TRUE))
))
names(domain_flags) <- names(domain_patterns)

matched_domains_vec <- apply(domain_flags, 1, function(row) {
  m <- names(domain_patterns)[as.logical(row)]
  if (length(m)==0) "No keyword match" else paste(m, collapse="; ")
})

grants_domains <- grants |>
  bind_cols(domain_flags) |>
  mutate(
    n_domain_matches   = rowSums(across(all_of(names(domain_patterns))), na.rm=TRUE),
    matched_domains    = matched_domains_vec,
    domain_match_group = case_when(
      n_domain_matches == 0 ~ "No keyword match",
      n_domain_matches == 1 ~ "One domain match",
      TRUE                  ~ "Multiple domain matches")
  )
```

```{r domain-volume, fig.width=11, fig.height=8}
purrr::map_dfr(names(domain_patterns), function(dn) {
  tibble(domain=dn,
         Posted     = sum(grants_domains$status_group=="Posted"     & safe_false(grants_domains[[dn]])),
         Forecasted = sum(grants_domains$status_group=="Forecasted" & safe_false(grants_domains[[dn]])))
}) |>
  pivot_longer(c(Posted,Forecasted), names_to="status_group", values_to="n") |>
  mutate(domain=fct_reorder(domain, n, .fun=sum)) |>
  ggplot(aes(x=n, y=domain, fill=status_group)) +
  geom_col(position="stack", width=.7) +
  geom_text(data=~group_by(.,domain)|>summarise(total=sum(n),.groups="drop"),
            aes(x=total,y=domain,label=comma(total)),
            inherit.aes=FALSE, hjust=-.12, size=3.4) +
  scale_fill_manual(values=c(Posted=pal$blue, Forecasted=pal$purple)) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.14))) +
  labs(title="Keyword Match Volume by NCDHHS-Relevant Domain",
       subtitle="An opportunity may match more than one domain",
       x="Keyword matches", y=NULL, fill=NULL) +
  theme(legend.position="top", panel.grid.major.y=element_blank())
```
# 15. County-Level Need: ACS Indicators

ACS 2019-2023 5-year estimates provide the demographic and socioeconomic baseline
for the need axis. County boundaries come from the Census TIGER/Line 2023 vintage
via the `tigris` package.

<div class="data-source-box">
**Source:** U.S. Census Bureau, American Community Survey 5-year estimates (2023 vintage).
Requires a free Census API key — set `census_api_key` in params or run
`tidycensus::census_api_key("YOUR_KEY")` before rendering.
**Access:** `tidycensus::get_acs()`
**Cached to:** `data/cache/api/acs_nc_county.rds`
</div>

```{r acs-download}
acs_cache <- file.path(cache_dir, "api", "acs_nc_county.rds")
nc_counties_cache <- file.path(cache_dir, "api", "nc_counties_tigris.rds")

# ACS variables — verified against 2023 5-year detailed tables
acs_vars <- c(
  total_pop        = "B01003_001",   # total population
  poverty_below    = "B17001_002",   # below poverty level
  poverty_universe = "B17001_001",   # poverty universe (denominator)
  median_income    = "B19013_001",   # median household income
  no_internet      = "B28002_013",   # no internet access (households)
  pop_under18      = "B09001_001",   # population under 18
  uninsured_male   = "B27001_006",   # male, under 6, uninsured (one component)
  disability_total = "B18101_001"    # civilian non-inst. population (disability base)
)

acs_county <- get_cached(acs_cache, function() {
  if (nchar(params$census_api_key) == 0) {
    message("No Census API key provided — skipping ACS download.")
    return(NULL)
  }
  tidycensus::get_acs(
    geography = "county", state = "NC",
    variables = acs_vars, year = 2023, survey = "acs5",
    output = "wide", progress_bar = FALSE
  ) |>
    janitor::clean_names() |>
    mutate(
      county_name   = str_remove(name, ", North Carolina"),
      poverty_rate  = poverty_below_e / pmax(poverty_universe_e, 1),
      med_income    = median_income_e,
      no_internet_n = no_internet_e,
      pop           = total_pop_e
    ) |>
    filter(!is.na(poverty_rate))
})

nc_counties_sf <- get_cached(nc_counties_cache, function() {
  tigris::counties(state = "NC", cb = TRUE, year = 2023, progress_bar = FALSE) |>
    sf::st_transform(4326)
})

acs_available <- !is.null(acs_county) && nrow(acs_county) > 0
geo_available <- !is.null(nc_counties_sf)
```

```{r acs-bar, fig.width=11, fig.height=7, eval=acs_available}
# Top 20 counties by poverty rate
acs_county |>
  slice_max(poverty_rate, n=20) |>
  mutate(county_name=fct_reorder(county_name, poverty_rate)) |>
  ggplot(aes(x=poverty_rate, y=county_name)) +
  geom_col(fill=pal$red, width=.72) +
  geom_text(aes(label=percent(poverty_rate, accuracy=0.1)), hjust=-.1, size=3.5) +
  geom_vline(xintercept=mean(acs_county$poverty_rate, na.rm=TRUE),
             linetype="dashed", color=pal$dark, linewidth=.8) +
  scale_x_continuous(labels=percent_format(), expand=expansion(mult=c(0,.15))) +
  labs(title="Top 20 NC Counties by Poverty Rate (ACS 2023)",
       subtitle="Dashed line = statewide average. Source: ACS 2019-2023 5-year.",
       x="Population below poverty level", y=NULL) +
  theme(panel.grid.major.y=element_blank())
```

```{r acs-map, fig.width=11, fig.height=7, eval=acs_available && geo_available}
map_dat <- nc_counties_sf |>
  left_join(acs_county |> select(geoid, poverty_rate, med_income),
            by=c("GEOID"="geoid"))

ggplot(map_dat) +
  geom_sf(aes(fill=poverty_rate), color="white", linewidth=.3) +
  scale_fill_distiller(palette="YlOrRd", direction=1,
                       labels=percent_format(accuracy=1),
                       na.value="gray90",
                       name="Poverty\nrate") +
  labs(title="NC County Poverty Rate (ACS 2023 5-year)",
       subtitle="Source: Census ACS, B17001. Model-derived; county estimates carry substantial uncertainty.",
       caption="Darker = higher poverty rate") +
  theme_void() +
  theme(legend.position="right",
        plot.title=element_text(face="bold", size=14),
        plot.subtitle=element_text(color=pal$gray))
```

```{r acs-unavailable, results='asis', eval=!acs_available}
cat('<div class="callout callout-warning">ACS data not yet downloaded.
Set a Census API key in params (<code>census_api_key</code>) and re-render,
or run <code>tidycensus::census_api_key("YOUR_KEY")</code> before rendering.
</div>')
```

# 16. Provider Shortage Areas: HRSA HPSA

HRSA Health Professional Shortage Areas (HPSAs) are the federal government's own
designation of inadequate delivery capacity. They are difficult to dispute in a
grant application narrative.

<div class="data-source-box">
**Source:** HRSA Bureau of Health Workforce — HPSA Find / Data Downloads.
**Access:** Direct CSV from `data.hrsa.gov`
**Disciplines downloaded:** Mental health, primary care
**Cached to:** `data/cache/api/hrsa_mh.rds`, `hrsa_pc.rds`
</div>

```{r hrsa-download}
hrsa_dl <- function(discipline_code, cache_name) {
  get_cached(file.path(cache_dir, "api", cache_name), function() {
    url <- paste0("https://data.hrsa.gov/DataDownload/DD_Files/",
                  "BCD_HPSA_FCT_DET_", discipline_code, ".csv")
    tryCatch(
      readr::read_csv(url, show_col_types=FALSE, progress=FALSE) |>
        janitor::clean_names(),
      error = function(e) { message("HRSA download failed: ", e$message); NULL }
    )
  })
}

hrsa_mh <- hrsa_dl("MH", "hrsa_mh.rds")
hrsa_pc <- hrsa_dl("PC", "hrsa_pc.rds")

process_hrsa <- function(df, discipline) {
  if (is.null(df)) return(NULL)
  df |>
    filter(str_to_upper(hpsa_status) == "DESIGNATED",
           str_to_upper(common_state_name) == "NORTH CAROLINA") |>
    mutate(
      discipline        = discipline,
      hpsa_score        = suppressWarnings(as.numeric(hpsa_score)),
      county_name       = str_remove(coalesce(hpsa_county_equivalent_name,""),
                                      " County| county"),
      shortage_type     = coalesce(hpsa_type_name, "Unknown"),
      rural_status      = coalesce(hpsa_rural_status_name, "Unknown")
    ) |>
    select(discipline, hpsa_id=hpsa_id, county_name, shortage_type,
           rural_status, hpsa_score,
           designation_date=hpsa_designation_date,
           withdrawal_date=hpsa_withdrawal_date)
}

hrsa_nc <- bind_rows(
  process_hrsa(hrsa_mh, "Mental health"),
  process_hrsa(hrsa_pc, "Primary care")
)
hrsa_available <- !is.null(hrsa_nc) && nrow(hrsa_nc) > 0
```

```{r hrsa-county-bar, fig.width=11, fig.height=7, eval=hrsa_available}
hrsa_county_summary <- hrsa_nc |>
  group_by(county_name, discipline) |>
  summarise(n_hpsa=n(), max_score=max(hpsa_score, na.rm=TRUE), .groups="drop")

hrsa_county_summary |>
  group_by(county_name) |>
  mutate(total=sum(n_hpsa)) |> ungroup() |>
  slice_max(total, n=40, with_ties=FALSE) |>
  mutate(county_name=fct_reorder(county_name, total)) |>
  ggplot(aes(x=n_hpsa, y=county_name, fill=discipline)) +
  geom_col(position="stack", width=.72) +
  scale_fill_manual(values=c("Mental health"=pal$purple, "Primary care"=pal$blue)) +
  scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.12))) +
  labs(title="HRSA Shortage Area Designations by NC County (top 40)",
       subtitle="Each row is one HPSA designation; a county may have several",
       x="HPSA designations", y=NULL, fill="Discipline") +
  theme(panel.grid.major.y=element_blank(), legend.position="top")
```

```{r hrsa-map, fig.width=11, fig.height=7, eval=hrsa_available && geo_available}
hrsa_map_dat <- hrsa_nc |>
  filter(discipline=="Mental health") |>
  count(county_name, name="mh_hpsa_count") |>
  mutate(county_name=str_squish(county_name))

nc_map_hrsa <- nc_counties_sf |>
  mutate(county_clean=str_remove(NAME," County| county")) |>
  left_join(hrsa_map_dat, by=c("county_clean"="county_name"))

ggplot(nc_map_hrsa) +
  geom_sf(aes(fill=mh_hpsa_count), color="white", linewidth=.3) +
  scale_fill_distiller(palette="Purples", direction=1,
                       na.value="gray90", name="MH HPSA\ndesignations") +
  labs(title="Mental Health HPSA Designations by NC County",
       subtitle="Source: HRSA HPSA data. Gray = no current designation or county boundary mismatch.") +
  theme_void() +
  theme(plot.title=element_text(face="bold",size=14),
        plot.subtitle=element_text(color=pal$gray))
```

```{r hrsa-table, eval=hrsa_available}
hrsa_nc |>
  arrange(discipline, desc(hpsa_score)) |>
  DT::datatable(rownames=FALSE, filter="top",
    options=list(pageLength=15, scrollX=TRUE, dom="Bfrtip",
                 buttons=c("copy","csv")),
    class="cell-border stripe hover")
```

```{r hrsa-unavailable, results='asis', eval=!hrsa_available}
cat('<div class="callout callout-warning">HRSA data could not be downloaded.
Check network access to <code>data.hrsa.gov</code> and re-render.</div>')
```

# 17. Local Health Outcomes: CDC PLACES

CDC PLACES provides model-based estimates of health outcomes at county, place,
tract, and ZCTA level. These are **modelled estimates**, not direct survey counts.

<div class="data-source-box">
**Source:** CDC PLACES 2025 release — County GIS-Friendly Format.
**Access:** CDC Open Data Portal (Socrata API), dataset ID `i46a-9kgh`
**Key measures used:** Frequent mental distress, depression prevalence
**Cached to:** `data/cache/api/places_nc_county.rds`
</div>

```{r places-download}
places_cache <- file.path(cache_dir, "api", "places_nc_county.rds")

places_nc <- get_cached(places_cache, function() {
  base_url <- "https://data.cdc.gov/resource/i46a-9kgh.csv"
  query    <- "?StateDesc=North+Carolina&$limit=200"
  tryCatch(
    readr::read_csv(paste0(base_url, query), show_col_types=FALSE, progress=FALSE) |>
      janitor::clean_names(),
    error=function(e) { message("PLACES download failed: ", e$message); NULL }
  )
})

places_available <- !is.null(places_nc) && nrow(places_nc) > 0

if (places_available) {
  # Select key mental-health-relevant measures
  key_measures <- c("MHLTH","DEPRESSION","BINGE","CSMOKING","ACCESS2","BPHIGH")
  places_long <- places_nc |>
    select(county_name=locationname, any_of(str_to_lower(key_measures)),
           matches("_crude(prev|value)")) |>
    pivot_longer(-county_name, names_to="measure", values_to="value") |>
    filter(!is.na(value), str_detect(measure, "crude"))
}
```

```{r places-mental-bar, fig.width=11, fig.height=7, eval=places_available}
# Mental distress or depression bar chart
mhlth_col <- names(places_nc)[str_detect(str_to_lower(names(places_nc)), "mhlth") &
                                str_detect(str_to_lower(names(places_nc)), "crude")][1]

if (!is.na(mhlth_col)) {
  places_nc |>
    rename(mhlth_val=all_of(mhlth_col)) |>
    mutate(mhlth_val=suppressWarnings(as.numeric(as.character(mhlth_val)))) |>
    filter(!is.na(mhlth_val)) |>
    slice_max(mhlth_val, n=20) |>
    mutate(county_name=fct_reorder(county_name, mhlth_val)) |>
    ggplot(aes(x=mhlth_val, y=county_name)) +
    geom_col(fill=pal$purple, width=.72) +
    geom_text(aes(label=paste0(round(mhlth_val,1),"%")), hjust=-.1, size=3.5) +
    scale_x_continuous(expand=expansion(mult=c(0,.15))) +
    labs(title="Top 20 NC Counties: Frequent Mental Distress Prevalence",
         subtitle="CDC PLACES 2025 model-based estimate. These are smoothed estimates, not direct observations.",
         x="% adults reporting frequent mental distress", y=NULL) +
    theme(panel.grid.major.y=element_blank())
} else {
  message("Mental distress column not found in PLACES data — check column names.")
}
```

```{r places-map, fig.width=11, fig.height=7, eval=places_available && geo_available && !is.na(mhlth_col)}
places_map <- nc_counties_sf |>
  left_join(places_nc |> rename(mhlth_val=all_of(mhlth_col)) |>
              mutate(mhlth_val=suppressWarnings(as.numeric(as.character(mhlth_val)))) |>
              select(locationname, mhlth_val),
            by=c("NAME"="locationname"))

ggplot(places_map) +
  geom_sf(aes(fill=mhlth_val), color="white", linewidth=.3) +
  scale_fill_distiller(palette="Purples", direction=1,
                       na.value="gray90", name="% frequent\nmental distress") +
  labs(title="Frequent Mental Distress Prevalence by NC County",
       subtitle="Source: CDC PLACES 2025. Model-based estimate.") +
  theme_void() +
  theme(plot.title=element_text(face="bold",size=14),
        plot.subtitle=element_text(color=pal$gray))
```

```{r places-unavailable, results='asis', eval=!places_available}
cat('<div class="callout callout-warning">CDC PLACES data could not be downloaded.
Check network access to <code>data.cdc.gov</code> and re-render.</div>')
```

# 18. Disaster Exposure: FEMA Declarations

Federal disaster declarations identify counties with accumulated disaster exposure.
Hurricane Helene (2024) and other recent events directly affect the disaster-recovery
program domain.

<div class="data-source-box">
**Source:** FEMA OpenFEMA Disaster Declarations Summaries v2.
**Access:** `https://www.fema.gov/api/open/v2/disasterDeclarationsSummaries`
**Scope:** North Carolina, declarations since 2000
**Cached to:** `data/cache/api/fema_nc.rds`
</div>

```{r fema-download}
fema_cache <- file.path(cache_dir, "api", "fema_nc.rds")

fema_nc <- get_cached(fema_cache, function() {
  url <- paste0(
    "https://www.fema.gov/api/open/v2/disasterDeclarationsSummaries",
    "?$filter=state%20eq%20'NC'%20and%20declarationDate%20ge%20'2000-01-01'",
    "&$orderby=declarationDate%20desc&$format=json&$top=2000"
  )
  resp <- tryCatch(
    httr2::request(url) |>
      httr2::req_timeout(30) |>
      httr2::req_error(is_error=function(r) FALSE) |>
      httr2::req_perform(),
    error=function(e) { message("FEMA request failed: ", e$message); NULL }
  )
  if (is.null(resp) || httr2::resp_status(resp) != 200) return(NULL)

  parsed <- httr2::resp_body_json(resp, simplifyVector=TRUE)
  if (is.null(parsed[["DisasterDeclarationsSummaries"]])) return(NULL)

  as_tibble(parsed[["DisasterDeclarationsSummaries"]]) |>
    janitor::clean_names() |>
    mutate(declaration_date=parse_date_safe(declaration_date),
           decl_year=year(declaration_date))
})

fema_available <- !is.null(fema_nc) && nrow(fema_nc) > 0
```

```{r fema-by-year, fig.width=11, fig.height=6, eval=fema_available}
fema_nc |>
  filter(!is.na(decl_year), decl_year >= 2000) |>
  count(decl_year, declaration_type) |>
  ggplot(aes(x=decl_year, y=n, fill=declaration_type)) +
  geom_col(width=.8, position="stack") +
  scale_fill_manual(values=c(DR=pal$red, EM=pal$orange, FM=pal$blue),
                    na.value=pal$gray) +
  scale_x_continuous(breaks=seq(2000,2030,2)) +
  scale_y_continuous(labels=comma, expand=expansion(mult=c(0,.12))) +
  labs(title="Federal Disaster Declarations in NC by Year",
       subtitle="DR = Major Disaster; EM = Emergency; FM = Fire Management",
       x=NULL, y="Declarations", fill="Type") +
  theme(panel.grid.major.x=element_blank(), legend.position="top",
        axis.text.x=element_text(angle=35, hjust=1))
```

```{r fema-county-map, fig.width=11, fig.height=7, eval=fema_available && geo_available}
fema_county <- fema_nc |>
  filter(declaration_type=="DR", !is.na(decl_year), decl_year >= 2015,
         !is.na(designated_area)) |>
  mutate(county_clean=str_remove(designated_area," \\(County\\)| County| county")) |>
  count(county_clean, name="dr_count")

nc_fema_map <- nc_counties_sf |>
  left_join(fema_county, by=c("NAME"="county_clean"))

ggplot(nc_fema_map) +
  geom_sf(aes(fill=dr_count), color="white", linewidth=.3) +
  scale_fill_distiller(palette="OrRd", direction=1,
                       na.value="gray90", name="Major disaster\ndeclarations\n(2015-present)") +
  labs(title="NC County Major Disaster Declarations (2015-present)",
       subtitle="Source: FEMA OpenFEMA v2. Includes Hurricane Helene 2024 and prior events.") +
  theme_void() +
  theme(plot.title=element_text(face="bold",size=14),
        plot.subtitle=element_text(color=pal$gray))
```

```{r fema-recent-table, eval=fema_available}
fema_nc |>
  filter(!is.na(decl_year)) |>
  arrange(desc(declaration_date)) |>
  slice_head(n=50) |>
  select(disaster_number, declaration_date, declaration_type,
         incident_type, designated_area, fy_declared=fy_declared) |>
  DT::datatable(rownames=FALSE, filter="top",
    options=list(pageLength=15, scrollX=TRUE, dom="Bfrtip",
                 buttons=c("copy","csv")),
    class="cell-border stripe hover")
```

```{r fema-unavailable, results='asis', eval=!fema_available}
cat('<div class="callout callout-warning">FEMA data could not be downloaded.
Check network access to <code>www.fema.gov</code> and re-render.</div>')
```

# 19. Federal Award History: USAspending Peer-State Comparison

For each CFDA program appearing in the grants snapshot, this section queries
USAspending to find which states have received awards in the last five years.
Programs where peer states (GA, VA, SC, TN) received funding but North Carolina
did not are the highest-priority gap candidates.

**Important caveat:** Award history reflects past applications, not eligibility.
A gap may mean NC never applied, applied unsuccessfully, or is ineligible by
practice but not by rule. Treat gaps as conversation starters, not proof of
missed money.

<div class="data-source-box">
**Source:** USAspending.gov API v2 — `spending_by_geography` endpoint.
**Access:** No authentication required.
**Scope:** Top 30 CFDAs by frequency in grants snapshot; FY2021-2025; grants and cooperative agreements.
**Cached to:** `data/cache/api/usa_cfda_*.rds`
</div>

```{r usaspending-download}
# Extract top CFDAs from grants snapshot
top_cfdas <- grants |>
  filter(!is.na(opportunity_assistance_listings),
         str_squish(opportunity_assistance_listings) != "") |>
  separate_rows(opportunity_assistance_listings, sep="\\s*;\\s*") |>
  mutate(cfda=str_extract(str_squish(opportunity_assistance_listings),
                           "^[0-9]+\\.[0-9]+")) |>
  filter(!is.na(cfda)) |>
  count(cfda, sort=TRUE) |>
  slice_head(n=30)

# Fetch spending by state for one CFDA
fetch_cfda_states <- function(cfda_num) {
  cache_path <- file.path(cache_dir, "api",
                           paste0("usa_", gsub("\\.","_",cfda_num), ".rds"))
  get_cached(cache_path, function() {
    body <- list(
      scope="place_of_performance", geo_layer="state",
      filters=list(
        award_type_codes=list("02","03","04","05"),
        time_period=list(list(start_date="2021-01-01", end_date="2025-12-31")),
        program_numbers=list(cfda_num)))
    resp <- tryCatch(
      httr2::request("https://api.usaspending.gov/api/v2/search/spending_by_geography/") |>
        httr2::req_body_json(body) |>
        httr2::req_timeout(25) |>
        httr2::req_error(is_error=function(r) FALSE) |>
        httr2::req_perform(),
      error=function(e) NULL)
    if (is.null(resp) || httr2::resp_status(resp) != 200) return(NULL)
    parsed <- httr2::resp_body_json(resp, simplifyVector=TRUE)
    if (is.null(parsed$results) || length(parsed$results)==0) return(NULL)
    as_tibble(parsed$results) |>
      transmute(cfda=cfda_num,
                state_code=shape_code,
                state_name=display_name,
                total_awards=aggregated_amount)
  })
}

# Fetch with a small delay between calls to respect rate limits
usa_results <- purrr::map(top_cfdas$cfda, function(c) {
  Sys.sleep(.25)
  fetch_cfda_states(c)
})

usa_df <- bind_rows(Filter(Negate(is.null), usa_results))
usa_available <- nrow(usa_df) > 0
```

```{r usaspending-gap-analysis, eval=usa_available}
focus_states <- c("NC","GA","VA","SC","TN")

usa_pivot <- usa_df |>
  filter(state_code %in% focus_states) |>
  mutate(received=total_awards > 0) |>
  select(cfda, state_code, received) |>
  pivot_wider(names_from=state_code, values_from=received, values_fill=FALSE)

# Gaps: peer received, NC did not
nc_gaps <- usa_pivot |>
  filter(!coalesce(NC, FALSE)) |>
  rowwise() |>
  mutate(n_peers_received=sum(c_across(any_of(c("GA","VA","SC","TN"))), na.rm=TRUE)) |>
  ungroup() |>
  filter(n_peers_received >= 1) |>
  arrange(desc(n_peers_received)) |>
  left_join(top_cfdas, by="cfda") |>
  rename(snapshot_count=n)
```

```{r usaspending-heatmap, fig.width=11, fig.height=8, eval=usa_available}
heat_dat <- usa_df |>
  filter(state_code %in% focus_states) |>
  mutate(
    has_awards=if_else(total_awards > 0, "Received", "Not received"),
    cfda=factor(cfda, levels=rev(sort(unique(cfda))))
  )

ggplot(heat_dat, aes(x=state_code, y=cfda, fill=has_awards)) +
  geom_tile(color="white", linewidth=.4) +
  scale_fill_manual(values=c("Received"=pal$green, "Not received"=pal$light),
                    na.value=pal$light) +
  labs(title="Federal Award Receipt by State and CFDA Program (FY 2021-2025)",
       subtitle="Green = at least one award received. Top 30 CFDAs by frequency in snapshot.",
       x="State", y="CFDA number", fill=NULL) +
  theme(panel.grid=element_blank(), legend.position="top",
        axis.text.y=element_text(size=8))
```

```{r usaspending-gap-table, eval=usa_available && nrow(nc_gaps) > 0}
nc_gaps |>
  select(cfda, n_peers_received, GA, VA, SC, TN, snapshot_count) |>
  mutate(across(c(GA,VA,SC,TN), ~ if_else(.x, "Yes", "No"))) |>
  DT::datatable(rownames=FALSE, filter="top",
    colnames=c("CFDA","Peer states w/ award","GA","VA","SC","TN","In snapshot"),
    options=list(pageLength=15, dom="Bfrtip", buttons=c("copy","csv")),
    class="cell-border stripe hover")
```

```{r usaspending-unavailable, results='asis', eval=!usa_available}
cat('<div class="callout callout-warning">USAspending data could not be retrieved.
Check network access to <code>api.usaspending.gov</code> and re-render.
Individual CFDA results are cached once retrieved.</div>')
```

# 20. Behavioral Health Capacity: SAMHSA Facilities

SAMHSA's N-SUMHSS survey provides facility-level counts of substance-use and
mental-health treatment capacity. Missing facilities in this dataset should not be
interpreted as proof that no services exist — survey participation is voluntary.

<div class="data-source-box">
**Source:** SAMHSA National Survey of Substance Abuse Treatment Services (N-SUMHSS / NSSATS).
**Download:** Visit `samhsa.gov/data` > N-SUMHSS > Download most recent public-use file.
**Expected local path:** `data/raw/samhsa/nsumhss-latest.csv`
**Note:** No unauthenticated public bulk download API exists. File must be downloaded manually.
</div>

```{r samhsa-load}
samhsa_paths <- c(
  file.path(data_dir, "raw", "samhsa", "nsumhss-latest.csv"),
  file.path(data_dir, "raw", "samhsa", "nsumhss-2023.csv"),
  file.path(data_dir, "raw", "samhsa", "nsumhss-2022.csv")
)

samhsa_file <- Filter(file.exists, samhsa_paths)[1]
samhsa_available <- !is.na(samhsa_file) && length(samhsa_file) > 0

samhsa_nc <- if (samhsa_available) {
  tryCatch(
    readr::read_csv(samhsa_file, show_col_types=FALSE) |>
      janitor::clean_names() |>
      filter(str_to_upper(coalesce(state,"")) == "NC"),
    error=function(e) { samhsa_available <<- FALSE; NULL }
  )
} else NULL
```

```{r samhsa-bar, fig.width=11, fig.height=7, eval=samhsa_available && !is.null(samhsa_nc) && nrow(samhsa_nc)>0}
county_col <- names(samhsa_nc)[str_detect(str_to_lower(names(samhsa_nc)), "county")][1]
if (!is.na(county_col)) {
  samhsa_nc |>
    rename(county=all_of(county_col)) |>
    count(county, sort=TRUE) |>
    slice_head(n=20) |>
    mutate(county=fct_reorder(county,n)) |>
    ggplot(aes(x=n, y=county)) +
    geom_col(fill=pal$purple, width=.72) +
    geom_text(aes(label=comma(n)), hjust=-.1, size=3.5) +
    scale_x_continuous(labels=comma, expand=expansion(mult=c(0,.14))) +
    labs(title="SAMHSA Treatment Facilities by NC County (top 20)",
         subtitle="Source: SAMHSA N-SUMHSS. Survey-based; absence does not equal no service.",
         x="Facilities in dataset", y=NULL) +
    theme(panel.grid.major.y=element_blank())
}
```

```{r samhsa-unavailable, results='asis', eval=!samhsa_available}
cat('<div class="callout callout-warning">
SAMHSA N-SUMHSS file not found at any of the expected paths:
<code>data/raw/samhsa/nsumhss-latest.csv</code><br>
Download the most recent public-use file from
<a href="https://www.samhsa.gov/data/data-we-collect/n-sumhss-national-substance-use-and-mental-health-services-survey/datafiles" target="_blank">samhsa.gov/data</a>
and save it to that path, then re-render.
</div>')
```

# 21. Overdose Mortality: CDC WONDER

County-level drug overdose death rates require a manual query from CDC WONDER
because the system does not expose a public REST API. Small-county values are
suppressed by NCHS and must remain missing — do not recode suppressed values as zero.

<div class="data-source-box">
**Source:** CDC WONDER Underlying Cause of Death (ICD-10 codes X40-X44, X60-X64, X85, Y10-Y14).
**Download:** `wonder.cdc.gov` — Underlying Cause of Death 1999-2022 (or current year), group by county, NC only.
**Expected local path:** `data/raw/cdc_wonder/overdose-mortality-county.txt` or `.csv`
**Note:** CDC WONDER does not have a public REST API. Export the query result as a text file.
</div>

```{r wonder-load}
wonder_paths <- c(
  file.path(data_dir, "raw", "cdc_wonder", "overdose-mortality-county.txt"),
  file.path(data_dir, "raw", "cdc_wonder", "overdose-mortality-county.csv")
)
wonder_file     <- Filter(file.exists, wonder_paths)[1]
wonder_available <- !is.na(wonder_file) && length(wonder_file) > 0

wonder_nc <- if (wonder_available) {
  tryCatch({
    raw <- readr::read_tsv(wonder_file[1], show_col_types=FALSE, comment="---") |>
      janitor::clean_names()
    # Normalize column names across WONDER export variants
    if (!"county" %in% names(raw) && "county_1" %in% names(raw))
      raw <- rename(raw, county=county_1)
    raw |>
      filter(!is.na(county), county != "Total") |>
      mutate(
        county_name = str_remove(county, ", NC$|, North Carolina$"),
        deaths       = suppressWarnings(as.numeric(deaths)),
        crude_rate   = suppressWarnings(as.numeric(crude_rate))
      )
  }, error=function(e) { wonder_available <<- FALSE; NULL })
} else NULL
```

```{r wonder-map, fig.width=11, fig.height=7, eval=wonder_available && geo_available && !is.null(wonder_nc)}
wonder_map_dat <- nc_counties_sf |>
  left_join(wonder_nc |> select(county_name, crude_rate),
            by=c("NAME"="county_name"))

ggplot(wonder_map_dat) +
  geom_sf(aes(fill=crude_rate), color="white", linewidth=.3) +
  scale_fill_distiller(palette="Reds", direction=1,
                       na.value="gray90",
                       name="Crude rate\nper 100k") +
  labs(title="Drug Overdose Mortality Rate by NC County",
       subtitle="Source: CDC WONDER. Gray = suppressed (<10 deaths) or data not matched.") +
  theme_void() +
  theme(plot.title=element_text(face="bold",size=14),
        plot.subtitle=element_text(color=pal$gray))
```

```{r wonder-bar, fig.width=11, fig.height=7, eval=wonder_available && !is.null(wonder_nc)}
wonder_nc |>
  filter(!is.na(crude_rate)) |>
  slice_max(crude_rate, n=20) |>
  mutate(county_name=fct_reorder(county_name, crude_rate)) |>
  ggplot(aes(x=crude_rate, y=county_name)) +
  geom_col(fill=pal$red, width=.72) +
  geom_text(aes(label=round(crude_rate,1)), hjust=-.1, size=3.5) +
  scale_x_continuous(expand=expansion(mult=c(0,.14))) +
  labs(title="Top 20 NC Counties by Drug Overdose Crude Mortality Rate",
       subtitle="Source: CDC WONDER. Suppressed counties (< 10 deaths) excluded.",
       x="Crude rate per 100,000", y=NULL) +
  theme(panel.grid.major.y=element_blank())
```

```{r wonder-unavailable, results='asis', eval=!wonder_available}
cat('<div class="callout callout-warning">
CDC WONDER file not found. CDC WONDER does not expose a public API.<br>
<strong>To download:</strong>
<ol>
<li>Go to <a href="https://wonder.cdc.gov" target="_blank">wonder.cdc.gov</a></li>
<li>Select <em>Underlying Cause of Death 1999-2020</em></li>
<li>Group by: State (NC only), County</li>
<li>ICD-10 codes: X40-X44, X60-X64, X85, Y10-Y14</li>
<li>Export as Tab-delimited text and save to <code>data/raw/cdc_wonder/overdose-mortality-county.txt</code></li>
</ol>
</div>')
```
