---
title: "Solve-A-Thon 2026"
author: "Kent Lee"
date: "`r Sys.Date()`"
geometry: margin=0.5in
output:
  rmdformats::readthedown:
      self_contained: true
      thumbnails: true
      lightbox: true
      gallery: true
      highlight: kate
      toc_float: yes
      code_folding: show
      code_download: true
params:
  project_dir: "/Users/kentalee/Documents/KALProjects/UNCProjects/SolveAThon_2026"
  results_dir: "outs"
  data_dir: "data"
  cache_dir: "cache"
  grants_file: "grants-search-202608182008.csv"
---

<style>
#content {
  max-width: 100% !important;
  padding-left: 40px !important;
  padding-right: 40px !important;
}

mark {
  background-color: #E6E6FA;
  color: black;
  border: 1px solid #E6E6FA;
  border-radius: 5px;
  font-family: Menlo, monospace;
  font-size: 0.85em;
}

img, .ggplot2-responsive {
  width: 80% !important;
  max-width: 80% !important;
  height: auto !important;
}

h1, h2 {
  border-bottom: 1px solid #E6E6FA;
  padding-bottom: 6px;
  margin-top: 34px;
}

h3 {
  margin-top: 26px;
}

.dataTables_wrapper {
  margin-top: 18px;
  margin-bottom: 30px;
  font-size: 0.88em;
}

table.dataTable thead th {
  background-color: #F4F6FA;
}

pre {
  border-radius: 6px;
}

.figure, .ggplot2-responsive {
  margin-top: 14px;
  margin-bottom: 24px;
}

blockquote {
  border-left: 4px solid #4A90E2;
  background-color: #F8F9FA;
  padding: 10px 16px;
  color: #333333;
}
</style>

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  eval = TRUE,
  message = TRUE,
  warning = FALSE,
  fig.align = "center",
  dpi = 150,
  out.width = "100%"
)

project_dir <- params$project_dir
results_dir <- file.path(params$project_dir, params$results_dir)
data_dir <- file.path(params$project_dir, params$data_dir)
cache_dir <- file.path(data_dir, params$cache_dir)

knitr::opts_knit$set(root.dir = project_dir)
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

options(bitmapType = "cairo")
```

```{r libraries, include = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
  library(lubridate)
  library(stringr)
  library(janitor)
  library(ggrepel)
  library(purrr)
  library(DT)
  library(scales)
  library(crosstalk)
  library(plotly)
})
```


## Screening Parameters

All screening logic is controlled from the config chunk below. Setting any parameter
to `NULL` disables that restriction and lets every row through.

* `ref_date_input` is the date the shortlist is built for. `NULL` uses today's date.
* `min_runway_days` is the minimum number of days between the reference date and the deadline. `NULL` accepts any deadline, including postings with no close date.
* `target_agencies` limits results to the federal agencies that issue the grant. `NULL` keeps every agency.
* `required_applicant_types` requires at least one listed applicant type to appear on the posting. `NULL` keeps every applicant type.
* `include_forecasts` keeps forecasted postings alongside posted ones. `NULL` or `TRUE` keeps both.
* `exclude_research_mechanisms` removes investigator-initiated research grants. `NULL` or `FALSE` keeps them.
* `exclude_restricted_competitions` removes single-source and invitation-only postings. `NULL` or `FALSE` keeps them.
* `award_floor_usd` is the minimum award size worth pursuing. `NULL` applies no floor.

```{r config}
is_unset <- function(x) {
  is.null(x) || length(x) == 0L || all(is.na(x))
}

ref_date_input <- "2026-09-30"
min_runway_days <- 45
target_agencies <- c("DHHS")
required_applicant_types <- c("state_governments")
include_forecasts <- FALSE
exclude_research_mechanisms <- TRUE
exclude_restricted_competitions <- TRUE
award_floor_usd <- 250000

ref_date <- if (is_unset(ref_date_input)) Sys.Date() else ymd(ref_date_input)
earliest_close <- if (is_unset(min_runway_days)) NULL else ref_date + days(min_runway_days)
keep_forecasts <- is_unset(include_forecasts) || isTRUE(include_forecasts)
apply_research_rule <- isTRUE(exclude_research_mechanisms)
apply_restricted_rule <- isTRUE(exclude_restricted_competitions)

lbl_agency <- if (is_unset(target_agencies)) {
  "Any issuing agency"
} else {
  paste0("Issued by ", paste(target_agencies, collapse = " / "))
}

lbl_forecast <- if (keep_forecasts) "Posted or forecasted" else "Posted, not forecast"

lbl_runway <- if (is.null(earliest_close)) {
  "Any deadline"
} else {
  paste0("Closes on or after ", format(earliest_close, "%b %d, %Y"))
}

lbl_eligible <- if (is_unset(required_applicant_types)) {
  "Any applicant type"
} else {
  paste0(
    paste(str_replace_all(required_applicant_types, "_", " "), collapse = " or "),
    " eligible"
  )
}

lbl_research <- if (apply_research_rule) "Not a research mechanism" else "Research mechanisms kept"
lbl_restricted <- if (apply_restricted_rule) "Open competition" else "Restricted competitions kept"

research_agency_regex <- "^HHS-NIH"

research_number_regex <- "^(PA|PAR|PD|RFA)-"

research_activity_regex <- paste0(
  "\\b(",
  paste(
    c("R01", "R03", "R13", "R21", "R24", "R25", "R34", "R35", "R41", "R42", "R43", "R44",
      "U01", "U18", "U19", "U24", "U54", "UG3", "UH3", "UM1", "P01", "P20", "P30", "P50",
      "K01", "K08", "K23", "K99", "T03", "T32", "T42", "F31", "F32", "DDRIG"),
    collapse = "|"
  ),
  ")\\b"
)

research_category <- "science_technology_and_other_research_and_development"

restricted_regex <- regex(
  paste(
    c("only the following applicant",
      "only the following entit",
      "single source",
      "sole source",
      "invited applicant",
      "is eligible to apply for this single",
      "not a competitive"),
    collapse = "|"
  ),
  ignore_case = TRUE
)

pal <- list(
  primary   = "#4A90E2",
  secondary = "#7B68A6",
  accent    = "#FF6B6B",
  neutral   = "#B8C2CC",
  ink       = "#222222",
  light     = "#F8F9FA"
)

pal_flag <- c("Kept" = "#4A90E2", "Dropped" = "#FF6B6B")

theme_sat <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, color = pal$ink),
      plot.subtitle = element_text(color = "#555555", margin = margin(b = 10)),
      plot.caption = element_text(color = "#777777", size = base_size - 4, hjust = 0),
      panel.grid.minor = element_blank(),
      axis.text.y = element_text(face = "bold"),
      plot.margin = margin(10, 20, 10, 10)
    )
}

theme_set(theme_sat())

make_dt <- function(df, page_length = 10, caption = NULL) {
  display <- df |>
    mutate(
      across(where(is.list), ~ sapply(.x, paste, collapse = ", ")),
      across(
        where(is.character),
        ~ if_else(!is.na(.x) & nchar(.x) > 150, paste0(substr(.x, 1, 150), "..."), .x)
      )
    )

  headers <- names(display) |>
    str_replace_all("_", " ") |>
    str_to_title()

  DT::datatable(
    display,
    colnames = headers,
    caption = caption,
    filter = "top",
    rownames = FALSE,
    width = "100%",
    extensions = "FixedHeader",
    options = list(
      pageLength = page_length,
      scrollX = TRUE,
      fixedHeader = TRUE,
      autoWidth = FALSE,
      searchHighlight = TRUE,
      dom = "itfpt"
    ),
    class = "cell-border stripe hover"
  )
}
```

```{r load_data, include = FALSE}
grants_data <- read_csv(file.path(data_dir, params$grants_file))

date_cols <- grep("date$", names(grants_data), value = TRUE, ignore.case = TRUE)
bool_cols <- grep("^is_", names(grants_data), value = TRUE, ignore.case = TRUE)

grants_data <- grants_data |>
  mutate(
    across(all_of(date_cols), ymd),
    across(all_of(bool_cols), as.logical)
  )

clean_grants_data <- grants_data |>
  select(
    where(~ !all(is.na(.x))),
    -c(agency_email_address_description, agency_contact_description, additional_info_url_description)
  ) |>
  mutate(
    close_date = if_else(close_date == as.Date("2099-01-01"), as.Date(NA), close_date),
    funding_instruments = str_split(funding_instruments, "\\s*;\\s*") |> map(~ str_trim(.x)),
    funding_categories = str_split(funding_categories, "\\s*;\\s*") |> map(~ str_trim(.x)),
    applicant_types = str_split(applicant_types, "\\s*;\\s*") |> map(~ str_trim(.x)), 
    top_level_agency_name = case_match(
      top_level_agency_name,
      "Department of Health and Human Services"       ~ "DHHS",
      "U.S. National Science Foundation"              ~ "NSF",
      "Department of Defense"                         ~ "DOD",
      "Department of State"                           ~ "DOS",
      "Department of the Interior"                    ~ "DOI",
      "Department of Justice"                         ~ "DOJ",
      "Department of Agriculture"                     ~ "USDA",
      "Department of Transportation"                  ~ "DOT",
      "Department of Commerce"                        ~ "DOC",
      "Department of Housing and Urban Development"   ~ "HUD",
      "National Aeronautics and Space Administration" ~ "NASA",
      "Department of Energy"                          ~ "DOE",
      "Department of Labor"                           ~ "DOL",
      "Department of Homeland Security"               ~ "DHS",
      "Department of Education"                       ~ "ED",
      "National Endowment for the Arts"               ~ "NEA",
      "Small Business Administration"                 ~ "SBA",
      "Office of National Drug Control Policy"        ~ "ONDCP",
      "Department of Veterans Affairs"                ~ "VA",
      "Department of the Treasury"                    ~ "TREAS",
      "Environmental Protection Agency"               ~ "EPA",
      "National Endowment for the Humanities"         ~ "NEH",
      "Department of Energy - Office of Science"      ~ "DOE-SC",
      "Millennium Challenge Corporation"              ~ "MCC",
      "Library of Congress"                           ~ "LOC",
      .default = top_level_agency_name
    )
  )

if (!"applicant_eligibility_description" %in% names(clean_grants_data)) {
  clean_grants_data$applicant_eligibility_description <- NA_character_
}
```

# Part 1. All Federal Grant Opportunities

```{r format_summarize, include = FALSE}
total_grants <- nrow(clean_grants_data)
unique_agencies <- n_distinct(clean_grants_data$top_level_agency_name)

if ("award_ceiling" %in% names(clean_grants_data)) {
  max_grant <- clean_grants_data |> 
    filter(!is.na(award_ceiling)) |> 
    slice_max(award_ceiling, n = 1, with_ties = FALSE)
  
  max_amount <- dollar(max_grant$award_ceiling)
  max_agency <- max_grant$top_level_agency_name
} else {
  max_amount <- "N/A"
  max_agency <- "N/A"
}
```

## Executive Summary

Welcome to the **Solve-A-Thon 2026** data portfolio tracker. This workspace isolates key operational trajectories across federal grant systems to guide team strategy.

* **Identified Opportunities:** Tracking <mark>`r total_grants`</mark> unique funding programs.
* **Sponsoring Bodies:** Distributed across <mark>`r unique_agencies`</mark> key federal agencies.
* **Top Ceiling Value:** The largest cataloged single award opportunity scales up to **`r max_amount`**, managed by the **`r max_agency`**.

---

## Visual Analysis

### Portfolio Volume by Sponsoring Agency
The figure below sorts federal agencies by the total count of funding announcements in our dataset, illustrating where investment pipelines are most heavily concentrated.

```{r agency_chart, fig.width=10, fig.height=6}
agency_summary <- clean_grants_data |> 
  count(top_level_agency_name, sort = TRUE) |> 
  head(15)

ggplot(agency_summary, aes(x = reorder(top_level_agency_name, n), y = n)) +
  geom_col(fill = pal$primary, width = 0.7) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5, color = "#333333") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "Top 15 Agencies by Grant Listing Volume",
    subtitle = "Posted and forecasted announcements in the August 18, 2026 snapshot",
    x = NULL,
    y = "Number of Grant Opportunities"
  ) +
  theme(panel.grid.major.y = element_blank())
```

<br>

### Financial Distribution of Award Ceilings
This map details the financial spreads across the highest-volume funding agencies, charting maximum award ranges using a relative logarithmic view.

```{r financial_chart, fig.width=10, fig.height=6}
if ("award_ceiling" %in% names(clean_grants_data)) {
  financial_data <- clean_grants_data |> 
    filter(!is.na(award_ceiling) & award_ceiling > 0) |> 
    filter(top_level_agency_name %in% agency_summary$top_level_agency_name[1:8])

  ggplot(financial_data, aes(x = top_level_agency_name, y = award_ceiling)) +
    geom_boxplot(outlier.color = pal$accent, fill = pal$light, color = pal$primary) +
    scale_y_log10(labels = label_dollar(scale_cut = cut_short_scale())) +
    labs(
      title = "Award Ceiling Variations by Core Agency Portfolio",
      subtitle = "Logarithmic scaling highlighting overall award funding distributions",
      x = "Agency Identifier",
      y = "Programmatic Award Ceiling ($)"
    ) +
    coord_flip()
} else {
  cat("Financial column 'award_ceiling' not detected in current input data frame structure.")
}
```

---

## Interactive Database Explorer

```{r interactive_table}
cols_to_remove <- c(
  "opportunity_id", 
  "opportunity_number", 
  "close_date_description", 
  "opportunity_assistance_listings", 
  "applicant_eligibility_description", 
  "forecasted_close_date_description", 
  "fiscal_year"
)

table_display <- clean_grants_data |>
  select(top_level_agency_name, everything()) |> 
  select(-any_of(cols_to_remove)) |>
  mutate(
    award_ceiling = if_else(!is.na(award_ceiling), scales::dollar(award_ceiling), "N/A"),
    award_floor = if_else(!is.na(award_floor), scales::dollar(award_floor), "N/A")
  )

make_dt(table_display, caption = "All opportunities in the source file.")
```